Realistic coupling enables flexible macroscopic traveling waves in the mouse cortex

Version 1

Abstract

Abstract

Traveling waves are ubiquitous in neuronal systems across different spatial scales. While microscopic and mesoscopic waves are relatively well studied, the mechanisms underlying the emergence of macroscopic traveling waves remain less understood. Here, by modeling the mouse cortex using spatial transcriptomic and connectivity data, we show that realistic cortical connectivity can generate a significantly higher level of macroscopic traveling waves than local and uniform connectivity. By quantifying the traveling waves in the 3-D domain, we discovered that the level of macroscopic traveling waves depends not only on the network connectivity but also non-monotonically depends on the coupling strength between neurons in the network. We also found that slow oscillations (0.5 - 4 Hz) are more likely to form large-scale, macroscopic traveling waves than other faster oscillations in the network with realistic connectivity. Together, our work shows how flexible macroscopic traveling waves can emerge in the mouse cortex and offers a computational framework to further study traveling waves in the mouse brain at the single-cell level.

Introduction

Traveling waves have emerged as a central topic among different large-scale spatiotemporal dy-namics observed in neuronal systems Muller et al. (2018). These waves are observed at different frequencies during various brain activities, ranging from slow-wave activity Massimini (2004); Liang et al. (2021), sleep spindles Muller et al. (2016), to faster oscillations in alpha Zhang et al. (2018); van Kerkoerle et al. (2014), beta Bhattacharya et al. (2022); Rubino et al. (2006); Davis et al. (2020), and gamma van Kerkoerle et al. (2014); Aggarwal et al. (2022) frequency bands. These waves also appear in various brain regions across species, including the hippocampus Lubenov and Siapas (2009); Patel et al. (2012); Zhang and Jacobs (2015), visual cortex Sato et al. (2012), prefrontal cor-tex Bhattacharya et al. (2022), motor cortex Rubino et al. (2006), or span across multiple brain regions Liang et al. (2023). More importantly, these waves have long been conjectured to regulate brain functions through mechanisms such as synaptic plasticity Ermentrout and Kleinfeld (2001); Muller et al. (2018). Indeed, researchers have recently found that traveling waves can modulate important brain functions such as memory, visual processing, and sleep Mohan et al. (2024); Aggarwal et al. (2022); Miyamoto et al. (2017).

The prevalence and important implications of traveling waves have prompted researchers to investigate their causes and organization. In these studies, computational modeling has been use-ful due to its flexibility. Researchers have developed both microscopic models Davis et al. (2020); Naze et al. (2015); Ermentrout and Kleinfeld (2001) and mesoscopic models such as neural mass models Roberts et al. (2019); Liang et al. (2023) to explore the mechanical properties and computational implications of traveling waves. However, most of these studies focus on microscopic or mesoscopic waves that appear in a single brain region. Large-scale, macroscopic waves across multiple brain regions have been less studied. Moreover, these models are usually constructed with local or uniform connectivity and lack the realistic geometry and connectivity of the brain. It has been shown that in a network of coupled non-linear oscillators, certain dynamics only arise when the number of oscillators is high enough, where local and global activities can coexist in the network Budzinski et al. (2022). At the same time, it has been argued that modeling neuronal networks at single-cell resolution will yield more precise discoveries than the mean-field model Deco et al. (2008). Therefore, it is ideal to study traveling waves in a large domain such as the cortex at single-cell resolution.

However, there are two significant challenges: First, a microscopic wiring diagram of the mouse brain at the whole-brain level is not available. However, we have mesoscopic connectivity data provided by the Allen Brain Atlas Oh et al. (2014); Knox et al. (2018); Harris et al. (2019), which will be used in this work. Second, simulating and measuring wave activity in a large 3D computational domain with realistic brain geometry is computationally expensive. Simulations of such scale are usually conducted on supercomputers Yamaura et al. (2020); Markram et al. (2015), and there are no established methods to quantify wave patterns on a large 3D domain.

In this work, we overcome these two challenges in order to study the emergence and propagation of macroscopic traveling waves in the mouse cortex. First, we propose an algorithm that integrates a spatial transcriptomic dataset that provides the detailed spatial and molecular profile of neurons Zhang et al. (2023), with voxelized connectivity data Knox et al. (2018) of the mouse brain. We use this algorithm to create neuron-to-neuron connectivity for one hemisphere of the mouse cortex, which we call Allen connectivity. We then constructed a 3D computational model with both excitatory and inhibitory neurons on top of the Allen connectivity. To overcome the second difficulty, we built our own parallel computational framework that can simulate the constructed model with a desktop GPU. We also extend the previously established method designed for 2-D data to our 3-D data to quantify the wave activities observed in simulations.

In our simulations, we found that the realistic Allen connectivity is capable of generating macroscopic waves across the mouse cortex with realistic propagating directions. Further quantification shows that Allen connectivity can generate a higher level of macroscopic wave activities than local and uniform connectivity. Moreover, we discover that the coupling strength between neurons non-monotonically affects the level of macroscopic wave activities that can emerge from different connectivity. Finally, we show that the realistic Allen connectivity is capable of generating a higher level of macroscopic traveling waves with different frequencies, especially in the Delta band (0.5 - 4 Hz), than uniform and local connectivity. Our work shows how realistic connectivity can enable flexible macroscopic waves across the mouse cortex. We also provide a computational framework to further study such traveling waves in the mouse brain at single-cell resolution.

Results

Construction of the cortical connectivity

A recent spatial transcriptomic dataset (Zhuang-ABCA1) Zhang et al. (2023) provides detailed positions of neurons in the mouse brain along with their molecular properties, including the neurotransmitter information (Fig. 1a, Use of the spatial transcriptomic data). Our goal is to connect neurons in Zhuang-ABCA1 based on realistic connectivity data of the mouse brain that has been experimentally measured. Previously, Oh et al. (2014) and Harris et al. (2019) have provided a mesoscopic connectivity matrix, known as the “projectome”, of the mouse brain. The projectome is derived from thousands of viral tracing experiments carried out within the mouse brain, and it measures the strength of connections between different brain regions at a regional level. This pro-Consider constructing a connectivity for N neurons that are contained in M voxels, and let Xi represent the physical position of the ith neuron. We have access to the voxelized projection matrix W, where Wab denotes the projection strength from voxel a to voxel b. Our connectivity algorithm requires only a single density parameter, ρ, which determines the average number of connections per neuron, thereby controlling the network’s sparsity (Use of the voxelized connectivity data). Using this information, we propose the following algorithm to construct the connectivity for the target neurons (Fig. 1b and Methods):

Construction of the cortical connectivity.

(a) Visualization of all neurons in the hemispherical dataset we use to build the cortical connectivity. The hemisphere contains about 1 million neurons in total, including about 300,000 cortical neurons. (b) Algorithm of building the connectome: i) First, we identify neuronal excitatory/inhibitory information based on the spatial transcriptomic data. ii) We then identify the voxelized position for all the neurons. iii) Based on the voxelized projection data, we calculate the number of connections between every pair of voxels (see Use of the voxelized connectivity data). iv) Connections are created randomly between the neurons in each pair of voxels. v) Repeat for all voxels in regions of interest. vi) Complete the connectivity. (c) The connectivity matrix shows the number of connections between different cortical regions(in log 10 scale). A full version of the connectivity matrix with annotated regions is shown in Figure 1—figure supplement 1. (d) Visualization of outgoing connections from sampled neurons in the primary visual cortex (VISp, left), primary auditory cortex (AUDp, middle), and the ventral anterior cingulate areas (ACAv, right).

Construction of cortical connectivity

Applying this algorithm to all cortical neurons identified in Zhuang-ABCA1, we obtain a cortical connectivity that consists of 297,812 neurons, and each neuron has 300 connections on average, resulting in a network sparsity of around 0.1 % (Construction and storage of the connectivity). The cortical connectivity we constructed is highly heterogeneous, where the number of connections between different regions can vary across four orders of magnitude (Fig. 1c). At the same time, both the in-degree and out-degree distribution of the connectome follow a realistic log-normal distribution Buzsáki and Mizuseki (2014) (Figure 1—figure supplement 1). Since the complete connectome is difficult to visualize, we show some connections from different cortical regions, including the primary visual, auditory cortex, and the ventral anterior cingulate area (Fig. 1d).

Network model and simulation

To simulate the Allen connectivity we just constructed, we now develop a computational model of the mouse cortex for computer simulation. The model incorporates both excitatory glutamatergic (AMPA) and inhibitory GABAergic neurons, with excitatory and inhibitory information provided by the original dataset (Zhuang-ABCA1). Both neurons are modeled as single-compartment, Hodgkin-Huxley type of neuron, whose firing behavior has been extensively described in Pospischil et al. (2008). Consequently, our network model contains two kinds of synapses (GABA and AMPA), both described by a first-order kinetic model (Neuronal and synaptic model).

Now, we need a mechanism to generate traveling waves in the simulation. According to previous studies, traveling waves can emerge in the mouse cortex both after sensory stimuli Mohajerani et al. (2013) or during spontaneous activities Liang et al. (2023). Here, we electrically stimulate all layer 4 neurons with an applied current as a way to simulate subcortical input to the cortex, since layer 4 is known to be the input layer Rockland (2019). Different levels of stimuli can be given by adjusting the magnitude of the applied current. Each simulation lasts for 3 seconds, during which the neurons are constantly under stimulation. To make such large-scale simulations possible on realistic time frames, we construct our own computational engine utilizing GPUs for simulation and visualization (Simulation).

Macroscopic traveling waves arise from realistic connectivity

Simulating the model with Allen connectivity, we immediately observe macroscopic traveling waves emerge in the network caused by the electrical stimuli to layer 4 neurons. In Figure 2 and Videos, we present a simulation in which a rhythmic oscillation occurs. The oscillation itself is well synchronized across the cortex, as shown by calculating the average voltage across the cortex and the raster plots of 40,000 sampled neurons (Fig. 2a). Next, we properly choose the time window to visualize the voltage activities of all the neurons in the network. We first visualize the intracellular voltage of all the individual neurons in the network (Fig. S2), but since the number of neurons are large, we proceed to visualize the average of the intracellular voltage of 300 neurons nearby at each location (Fig. 2b and d). This reveals an apparent macroscopic traveling wave in the anterior-posterior direction through the cortex, which has been previously reported Massimini (2004); Liang et al. (2021); Aggarwal et al. (2024). We also observe the anterior-posterior macroscopic wave in both directions (Fig. 2c and 2e). In fact, the two waves can propagate in the reverse direction right after each other, indicating a possible feedforward-feedback mechanism.

Anterior-posterior macroscopic traveling waves emerge from Allen connectivity

(a) Top: The average intracellular voltage of all neurons; bottom: Raster plots for 40000 neurons, which are evenly sampled from different cortical regions. The red and blue dashed lines indicate the time windows that are visualized in (b-e). (b) Visualization of average intracellular voltage across different locations in the cortex during the time window indicated by the red dashed lines in (a). (c) Visualization of the macroscopic traveling wave by coloring the neurons based on the first time they fire during the same time window visualized in (b). (d-e) Same plot as (b-c), but they visualize a different time window of the same simulation, indicated by the blue dashed lines in (a). The direction of the traveling wave is opposite to the one shown in the previous figures. See Videos for the movie of the simulation.

Figure 2—figure supplement 1. Visualization of the macroscopic traveling waves at single-neuron resolution. (a-b) Visualization of the same

Higher level of macroscopic wave emerges from realistic connectivity

In past modeling studies of traveling waves, it is common to adopt uniform-random, local, or a combination of the two kinds of connectivity Liang et al. (2023). Therefore, we now simulate the same model with different connectivity in order to compare and investigate whether such macroscopic waves that are presented in Figure 2 only emerge from the Allen connectivity. For comparison, we constructed two additional types of connectivity that are common in past studies of traveling waves:

Uniform connectivity: Neurons are connected across the entire cortex randomly (Fig. 3b).

Local connectivity: Neurons are connected to a fixed number of their closest neighbors (Fig. 3c).

Top: schematics of Allen (a), local (b), and uniform connectivity (c); bottom: Visualization of outgoing connections from the same 50 randomly selected neurons in the primary visual cortex (VISp) for the three connectivity. Figure 3—figure supplement 1. Comparison of the three connectivity matrices: a Allen, b local and c uniform connectivity. For local connectivity, the diagonal terms represent the connections within the same region. And the off-diagonal terms exist because these regions are physically adjacent to each other.

Here we visualize the outgoing connections from 50 randomly selected neurons in the primary visual cortex to show the difference between the three connectivity. Under Allen connectivity, the outgoing connections project to different regions heterogeneously (Fig. 3a); with local connectivity, the outgoing connections only reach around the adjacent areas where the sampled neurons are located (Fig. 3b); and with uniform connectivity, the outgoing connections tend to cover the whole cortex evenly (Fig. 3c). Although the topology of three connectivity is different, they share the same out-degree distribution based on the way they are constructed (see Construction and storage of the connectivity).

To properly quantify the traveling wave observed in simulations, we applied a phase-based method from image processing Fleet and Jepson (1990), which has been adopted to analyze neuronal dynamics Rubino et al. (2006), to our simulation data. First, we performed the Hilbert transform on the time series of the intracellular voltage signal for all neurons (Fig. 4a) to transform the voltage signal into a phase signal. Since neurons are spatially distributed according to a realistic dataset rather than arranged on a regular grid required for further calculations, we constructed a Cartesian grid defined by the Allen Brain Atlas at a resolution of 100 µm and interpolate the phase signal of neurons onto the grid points (Fig. 4b) to obtain the phase field. We then compute the gradient field of the phase field on the 3D grid (Fig. 4c). This allows us to calculate the phase gradient directionality (PGD), an indicator for planar wave activity Rubino et al. (2006), which will be used to measure the level of macroscopic wave activities in our simulations (see Quantitative measurement of neuronal activity).

Schematics of analyzing 3D spatiotemporal data using phase gradient.

(a) A single frame from a simulation with Allen connectivity, where neurons are colored based on their intracellular voltage. (b) The same frame as in (a), but with voltage data transformed into a generalized phase and interpolated onto a regular 3D grid. (c) Calculation of phase gradient field based on the phase field presented in (b). The arrows indicate the gradient directions at different locations. Please see Quantitative measurement of neuronal activity for details of the calculation presented in this figure.

In Figure 5 and Videos, we compare the PGD produced by three representative simulations with different connectivity but the same parameters otherwise. We compare three time windows of the same length (70ms) where PGD reaches its maximum during the corresponding simulations. Although both Allen and local connectivity produce visible wave activity, Allen connectivity produces waves that are much more macroscopic, while local connectivity is capable of producing locally organized planar waves, which are the most frequently form of waves observed in previous expeimental studies Lubenov and Siapas (2009); Bhattacharya et al. (2022) and computational models with local connectivity Keane and Gong (2015). However, uniform connectivity did not generate any apparent structure. This qualitative observation was confirmed by the PGD calculations: uniform connectivity exhibited correspondingly lowest PGD, whereas Allen connectivity resulted in significantly higher phase gradient directionality (Fig. 5d).

Allen connectivity produces a higher level of macroscopic wave activity than local and uniform connectivity.

(a-c) Time sequences of three simulations with Allen (a), local (b) and uniform (c) connectivity under the same stimulus. Colors represent the generalized phase of each neuron, which is further normalized for each simulation between 0 and 1. (d) Phase gradient directionality (PGD) of the corresponding simulations with respect to time. The three time windows are chosen such that the maximum PGD is reached. Please see Video 2 for animation of the three simulations.

Dynamic regime of the network affects macroscopic traveling waves

Traveling waves are known to propagate differently in a network under different states, depending on the excitability of the neurons and the strength of connections between them, even if the connectivity remains the same Muller et al. (2018). Here, after comparing the wave activities that emerge in three different connectivity for one parameter setting, we proceed to examine whether the Allen connectivity always generates more macroscopic wave activities under different conditions. To do so, we systematically alternate the coupling strength of the network and the level of stimuli, while fixing other parameters for each connectivity in each simulation (Simulation).

As a result, we observe that various kinds of oscillations arise from different connectivity under different conditions. As presented in Figure 6 and Videos, when the network is weakly coupled, all three connectivity are able to produce a certain level of macroscopic wave activities, indicated by the PGD calculation (Fig. 7b). Still, Allen connectivity produces the highest level of wave activity in this coupling threshold. We also found the network to be well synchronized for both Allen and uniform connectivity, but not for the local connectivity (Fig. 6a-c). However, when the coupling strength doubles from this level, the network goes into an asynchronous state for both Allen and uniform connectivity (Fig. 6d-f). This asynchronous firing behavior also disrupts the macroscopic wave activity that arises previously in the weakly coupled network. Interestingly, the local connectivity is capable of retaining the same level of macroscopic wave activities, which are higher then the Allen and uniform connectivity for this coupling threshold (Fig. 7b). Furthermore, if we keep increasing the coupling strength of the network, the synchrony will return to a high level for Allen and uniform connectivity, consequently increasing the level of macroscopic wave activity in the network (Fig. 6g-i).

Different oscillations and spatiotemporal patterns emerge from the cortical network with different connectivity and coupling strength.

For each panel, the raster plot (top) shows the firing behavior of the population and the visualization (bottom) shows the intracellular voltage activity across the cortical surface during the period marked in red dashed lines on the corresponding raster plot. There are 9 panels in total, corresponding to three different connectivity and coupling strengths. Left to right: Allen, local and uniform connectivity; Top to bottom: weak (50% baseline), medium (baseline) and strong (150% baseline) coupling strength.

Macroscopic wave activity is strongly affected by the coupling strength of the network and the frequency of the oscillation.

(a-b) The average phase gradient directionality (PGD) calculated from simulations of different connectivity under different applied currents (a) or coupling strengths (b). The error bar represents the values from other simulations with different coupling strengths (a) or applied currents (b). (c) The average PGD calculated from simulations of different connectivity under different kinds of oscillations categorized by the dominant oscillation frequency (see Quantitative measurement of neuronal activity).

From these simulations, we found that the level of macroscopic wave activity is relatively insensitive to the magnitude of the stimuli: under the same coupling strength, changing the level of the stimuli does not strongly change the level of macroscopic wave activity. However, the coupling strength in the network can drastically change the level of macroscopic wave activity non-monotonically (Figure 7 and Figure 7—figure supplement 1). We found that the level of PGD produced with local connectivity is largely consistent across different dynamic regimes. However, Allen and uniform connectivity are much more sensitive to the change of coupling strength. For Allen connectivity, a high level of macroscopic wave activity can emerge at both low and high coupling strengths, but not during medium coupling strength, in which local connectivity creates the highest wave activities at this threshold, as can be seen from previous Figure 6. At the same time, the uniform connectivity consistently creates a lower level of wave activities across different coupling strengths, except at the weakest coupling strengths.

Slow oscillations are more likely to form macroscopic traveling waves

In the study of neural oscillations, researchers have found that different frequencies usually indicate different brain functions and behaviors Buzsáki and Draguhn (2004). As we mentioned in the introduction, traveling waves are observed across a wide range of frequencies, but it is unclear whether there is a relationship between the kind of oscillation and the scope of the traveling wave that can arise from the oscillation. Because we are able to induce a range of oscillations of different frequencies, as shown in Figure 6, these results now allow us to investigate the relationship between the frequency of the oscillation and its corresponding scope. For each simulation, we conduct spectral analysis on the average voltage of the whole population and find out the dominant frequency of the voltage oscillation based on the power density. For each frequency band (Delta: 0.5 - 4 Hz, Theta: 4 - 8 Hz, Alpha: 8 - 12 Hz, Beta: 12 - 30 Hz and Gamma: 30 - 100 Hz), we gather the simulations that produce the oscillation in the frequency band and calculate the average PGD across these simulations (Quantitative measurement of neuronal activity).

Surprisingly, from Fig. 7c, we found that the level of PGD is highly correlated with the frequency band of the oscillation. For slow oscillations in the Delta frequency band, the Allen connectivity is able to support a much higher level of macroscopic wave activity on average than uniform and local connectivity. However, for Theta oscillations, the level of PGD for Allen connectivity significantly drops to the same level that is comparable to local connectivity. For Alpha oscillations, the level of macroscopic wave activity rises again, and drops for faster oscillations in Beta and Gamma frequencies. If we compare across connectivity, local connectivity again supports a similar level of PGD across all oscillation frequencies, while Allen connectivity particularly favors the slow frequency. Uniform connectivity again produces the lowest level of macroscopic wave activity compared to the other two connectivity. These findings conjecture that slow oscillations are more likely to propagate as macroscopic traveling waves within the mouse cortex.

Discussion

Construction of the Allen connectivity

In this work, we present an algorithm that integrates the spatial transcriptomic dataset that provides the spatial distribution and molecular properties of neurons across the mouse brain, and the mesoscopic connectivity dataset that offers the connectivity information between brain regions on a voxelized level. The algorithm is efficient and can be easily adapted for similar datasets of the mouse brain or for even other species. In essence, our algorithm belongs to the family of community generating algorithms, in particular the stochastic block model Holland et al. (1983). The projectomme Knox et al. (2018) we use to build the Allen connectivity is derived from a series of viral-tracing experiments that potentially reflect the connection strength between different regions Oh et al. (2014). However, the relationship between the actual connectome and the mesoscopic projectome remains unclear. For instance, we did not account for variations in synaptic strength across different connections. Moreover, the projection data used is not cell-type specific. A more updated thalamocortical dataset Harris et al. (2019) exists but is only on the regional level and therefore was not used in this work. An important future direction will be to improve the algorithm to integrate more comprehensive data on different types of connectivity in the cortex.

Cortical model and simulation

Compared to previous large-scale models Capone et al. (2023); Roberts et al. (2019), our model advances in incorporating the detailed spatial distribution of neurons in the 3-D computational domain. Without this information, spatiotemporal dynamics cannot be accurately analyzed and visualized. Recent research Pang et al. (2023); Gulbinaite et al. (2024) has highlighted the significant role that brain geometry plays in neuronal functions. In this work, we did not further use the spatial information of neurons to model aspects such as the layered structure of the cortex, which would be an interesting direction for future work.

At the same time, although we are using one of the largest spatial transcriptomic datasets up to date, 300,000 neurons are still only a small portion of the total neuronal population in the mouse cortex. The number of neurons in a network will have important effects on aspects such as synaptic scaling. Among the total population, we only modeled two kinds of neurons, which is also a simplification. Therefore, enriching our model by including a higher number of different types of neurons and studying how they affect the conclusion in this paper is highly meaningful.

For simulations, we choose to stimulate the total population of layer 4 neurons as a way to generate traveling waves, which is quite rare under biologically normal conditions. We also did not include any subcortical structure, such as the thalamus, which is vital in cortical dynamics like slow-wave activity Steriade et al. (1993). In this work, we choose two parameters that affect the dynamic regime of the network, which are the level of the stimuli controlled by the applied current and the coupling strength between neurons. There are many other things that affect the dynamical regime of a network, for example, the excitatory and inhibitory balance is vital for different cortical functions Ahmadian and Miller (2021). Incorporating these perspectives is also important for future investigations.

Synaptic and conduction delays are essential to cortical dynamics such as traveling waves Petkoski and Jirsa (2022); Deco et al. (2009); Golomb and Ermentrou (2000). In our model, because the synapses are modeled by a first-order kinetic model (Neuronal and synaptic model), both excitatory and inhibitory synapses have a rising time that mimics the synaptic delay. However, we did not include conduction delay in our study. Conduction delay is thought to have a proportional relationship with the white matter path length of the connection between two neurons Lemaréchal et al. (2022). In our model, although we have the 3D positions of neurons, we do not have geometric information about the cortical manifold. Two neurons can be very close in the Cartesian coordinates measured by the Euclidean distance, but very far in terms of the length of the actual connection in the brain. Therefore, incorporating accurate conduction delay in the model is an important future direction.

Quantitative measurements of 3-D traveling waves

There are many techniques available for identifying and measuring large-scale neuronal spatiotemporal patterns Townsend and Gong (2018); Gutzen et al. (2024). However, most of these algorithms are designed to analyze 2-D data. Our simulation data is intrinsically 3-D, thus presenting new challenges for measurement. For instance, we need to identify neuron boundaries to exclude unrealistic boundary effects when calculating the phase gradient field (Quantitative measurement of neuronal activity). Additionally, working with high-resolution 3-D data is computationally expensive. Despite these challenges, we extended and verified the previous phase-based method to quantify the level of macroscopic wave activity in our simulations. However, as noted in the original work Rubino et al. (2006), phase gradient directionality only indicates planar wave activity and may fail to identify other wave types such as spiral waves, which are observed in the cortex Muller et al. (2016). We also did not analyze other cortical dynamics, such as sinks and sources, which have appeared in our simulation. Extending more sophisticated methods from 2-D to 3-D to identify diverse spatiotemporal patterns in experimental or simulated data will be a meaningful direction for future research.

Comparing with experimental data

Across the paper, we have chosen the intracellular voltage of neurons as the primary data for the analysis. The voltage signals, though do not directly resemble experimental measures such as local field potential (LFP), provide a quick way to analyze the overall dynamics of the whole network. There is strong experimental evidence of cortical spatiotemporal waves from measurements of LFP and multiunit activity reflecting extracellular electrical potentials in regions across the cortex Destexhe et al. (1999). Recent experimental studies of cortical wave patterns have achieved significant improvements in spatiotemporal resolution via methods such as wide-field calcium imaging Manley et al. (2024) and optical voltage imaging Liang et al. (2021). Our simulation framework allows many opportunities to compare with these experimental data.

However, one must be careful when comparing these simulations to experimental data, especially LFPs or other methods measuring extracellular potential rather than intracellular activity. Further work is needed to connect high-resolution mathematical models with experimental data based on the imaging techniques used. For example, Davis et al. (2020) has simulated LFP from a high-resolution leaky integrate-and-fire (LIF) network model to compare with real LFP recordings. In future work, we plan to develop new methods to compare model voltage traces with electro-physiological recordings.

Neural oscillation and traveling wave

Different neural oscillations are known to have different generating mechanisms. For example, the gamma oscillation is conjectured to be caused by coordinated interaction of excitatory and inhibitory neurons in the cortex Buzsáki and Wang (2012). And slow-wave activity is conjectured to be caused by thalamocortical interaction Steriade et al. (1993). In our work, we did not fully investigate how different kinds of oscillations emerge in the network, but only study how these oscillations produce different traveling waves in the cortex a posteriori. Through statistical observation, we found that the slow oscillations are more likely to form macroscopic traveling waves. In the future, it is essential to combine established understanding of mechanisms behind different oscillations with experiments to further verify the conclusions drawn in this work.

Methods and Materials

Use of the spatial transcriptomic data

The original spatial transcriptomic study Zhang et al. (2023) provides four datasets of cell information identified from the mouse brain. Among these, we selected the largest dataset, Zhuang-ABCA-1, which includes 2.8 million cells from one hemisphere, with approximately 2.6 million (N = 2,616,328) cells registered to the Allen Brain Atlas. Of these 2.6 million cells, about half are non-neuronal cells and were removed before constructing the connectivity. Since we are only modeling the glutamatergic and GABA-ergic neurons, we only kept these two kinds of neuronal populations, totaling approximately 1 million neurons (N = 1,050,440). As a result, our final dataset consists of N = 1,050,440 neurons (Fig. 1a) in one hemisphere. Within this population, the isocortex contains 297,812 neurons. The cortical connectivity is constructed by applying our algorithm to the total cortical population.

Use of the voxelized connectivity data

The projection data we use in this study comes from Knox et al. (2018), which provides the connection strength between any pairwise voxel in the isocortex. The most important part of using the projection data is determining the number of connections between neurons in any pairwise voxels based on the connection strength. Let W represent the voxelized connectivity matrix, where Wab represents the connection strength from voxel a to voxel b. In this work, we assume the number of connections between two voxels is proportional to the connection strength, meaning the number of connections from voxel a to voxel b is Nab = λWab, where λ is a fact that will be calculated based on the density parameter.

Before building the connectivity, we first choose the density parameter, ρ, which is the average number of connections per neuron. Let vector x represents the number of neurons in each voxel. If λ = 1, then the total number of connections in the whole connectivity is Ntotal = xT W x, and the average number of connections per neurons is , where M is the total number of neurons. Therefore, if we let , we will get the ideal average number of connections per neurons. Consequently,. In the code, after calculating beforehand, we can multiply the whole projection matrix W by λ to derive a connection matrix, whose each entry now represents the number of connections between any two voxels.

Construction and storage of the connectivity

Three types of connectivity are used in our simulations: Allen, local and uniform connectivity. Allen connectivity is constructed by applying our algorithm to the cortical neurons in Zhuang-ABCA1 based on the method described above. Uniform and local connectivity are constructed in a similar fashion: for each neuron, we first read out the number of outgoing connections in the corresponding Allen connectivity, then we randomly choose the same number of neurons from the total population without repetition for the uniform connectivity, or the same number of nearest neighbors for local connectivity. By this way, the out-degree distribution of all three connectivity will be the same, making them more comparable. In partiuclar, local connectivity is constructed through nearest-neighbor coupling, where we use the K-nearest neighbors (KNN) algorithm Cover and Hart (1967) to choose a fixed number of outgoing connections to nearest neighbors for each neuron based on spatial position. For all three connectivity, self-connection is not allowed.

When running our connectivity algorithm to build the Allen connectivity, a large number of neurons can pose some challenges to our connectivity algorithm. To tackle this challenge, we loop through the voxels for a significant speed-up. As a result, our algorithm will take O(MN) time, where M is the number of neurons that contain all the neurons and N is the number of voxels. N has a maximum since the computational domain is finite, and in the end, this algorithm is about two orders of magnitude faster than iterating through every neuron.

In our model, we choose ρ = 300, which means there are about 90 million synapses in the model since we have 297,812 neurons. To store such huge connectivity, we use a compressed row form representation of the connectivity matrix, which is also used in similar packages Dai et al. (2020).

Neuronal and synaptic model

We use a Hodgkin-Huxley type neuron model that can produce different firing behaviors of a cortical neuron Pospischil et al. (2008). Each neuron has four ionic channels: voltage-gated sodium, voltage-gated potassium, slow non-inactivating potassium, and leak channels. The pyramidal neurons will have all four ionic channels, while the inhibitory neurons do not have the slow non-inactivating potassium channels because they are fast-firing and do not usually exhibit firing adaptation. Detailed equations of each current and the change of gating variables are provided in the appendix. The parameters of the neuronal model are also drawn from the original paper and presented in Table S1 in the appendix.

For synapses, we model both glutamatergic (AMPA) or GABAergic synapses in our cortical network. Both synpases are modeled by a conducatance-based, first-order kinetic model. Please see the supplement for more details.

Simulation

In our system, we have to solve two sets of equations at every time step: the Hodgkin-Huxley equations for the electrophysiology model and the conductance-coupled synaptic equations. We solve the Hodgkin-Huxley equations using an established implicit leapfrog method DeWoskin et al. (2015), allowing for large timesteps due to increased stability. Both electrophysiology and synaptic equations are solved parallelly on GPU in our framework. The overall parallelism is based on the neuron, which means each set of Hodgkin-Huxley equations of a neuron is put up on an individual thread of GPU to solve at every timestep. Similarly, at any time step, we first check for all neurons firing and then assign each thread to process each of the firing neurons’ downstream connections. Please see the supplements for a detailed description of the numerical methods.

Each simulation lasts for 3 seconds, and we only record the last 1 second of the simulation to avoid the transient at the beginning of the simulation. The initial conditions of all neurons are randomly chosen at the beginning of each simulation. To be more specific, the initial voltage of all neurons is drawn from a uniform distribution between -65 mV and -5 mV, and all gate variables are drawn from a uniform distribution between 0.02 and 0.05. For each simulation, we collect the voltage traces of all neurons with a sampling frequency of 2000 Hz.

In all simulations, the total excitatory population of layer 4 neurons is electrically stimulated with an applied current. All layer 4 neurons receive the same level of applied current and all synapses have the same weight throughout each simulation. In order to adjust the level of stimuli, we change the applied current that ranges from 0.9 nA to 1.9 nA with a step of 0.1 nA. In the simulations, we also modify the coupling strength of the network, which is done by factoring the conductance of both excitatory and inhibitory synapses at the same time. The baseline conductance of an excitatory synapse is 30 nS, and we change the conductance within 15 to 50 nS with a step of 5 nS. The inhibitory conductance is always 10% of the excitaory conductance.

Quantitative measurement of neuronal activity

To better quantify spatiotemporal dynamics, we extend a phase-based method Rubino et al. (2006) that has primarily been used for 2-D data but naturally generalizes to 3-D. First, we apply the Hilbert transform to the voltage traces of all neurons and extract their phase, treating each neuron as a nonlinear oscillator. Next, we construct a regular 3-D Cartesian grid following the discretization of the Allen Brain Atlas common coordinate framework (ABA-ccf), resulting in a grid of size 132 ∗ 80 ∗ 114, with a spatial resolution of 100µm in each direction. We then linearly interpolate the generalized phase of all neurons to their adjacent grid points, effectively mapping neuronal phase information onto the 3-D grid.

Let ϕ(x, t) represent the interpolated phase field on the 3-D grid. To compute phase gradient directionality (PGD), we first calculate the phase field gradient, ∇ϕ, using a finite difference method. We then apply the following formula to calculate P, which represents the PGD: Note that the horizontal bar means taking the average across the whole spatial domain, and P is a function of time. Higher PGD means the gradient field at different spatial locations is aligned, and lower PGD means the opposite. In calculating the gradient field, the boundary cells that contain no neurons are excluded. This is done by using the PyVistas package Sullivan and Kaszynski (2019) by masking off the points that have no neurons to interpolate for. Both phase field interpolation and gradient computation are done by using the intrinsic functions in the 3-D Visualization Toolkit (VTK) Schroeder et al. (1998).

For spectral analysis of the oscillation, we first calculate the global average voltage by averaging the intracellular voltages of all neurons. We then apply a discrete Fourier transform to the average voltage trace with the signal.periodogram function from the SciPy package Virtanen et al. (2020). Given a sampling frequency of 2000 Hz, we compute the one-sided power spectral density for frequencies ranging from 0 to 1000 Hz and identify the frequency with the highest density, which we define as the dominant frequency. This value is used in plotting 7 and 2.

Visualization

Throughout the paper, we visualize neurons by coloring them based on their intracellular voltage. The visualization is straightforward for a single neuron such as in 1, but in 2, 5 and 6, we also need to calculate the average voltage of 300 neurons nearby. This is done by identifying the closest 300 neighbors for each neuron based on their spatial distributions and is calculated in post-processing. Additionally, Fig. 2c and e are visualized based on the time that each neuron fires for the first time under the chosen time window. For 6, the majority of the time window is chosen manually to fully show the spatiotemporal patterns of the corresponding simulation. All visualizations are made using the PyVistas package (Sullivan and Kaszynski (2019)).

eLife Assessment

This valuable work presents a novel computational framework for modeling macroscopic traveling waves in the mouse cortex by integrating open-source connectomic and transcriptomic data into a spiking network model. This approach allows the computational model to assign excitatory/inhibitory connections based on neurotransmitter profiles and extends simulations to the 3D domain. The authors present results that demonstrate how spatiotemporal dynamics such as slow oscillations (0.5-4 Hz) emerge and self-organize at the whole-brain scale. This study provides convincing initial insights into the structural basis of traveling waves at the whole-brain scale in the mouse.

Reviewer #1 (Public review):

Summary:

The manuscript "Realistic coupling enables flexible macroscopic traveling waves in the mouse cortex" by Sun, Forger, and colleagues presents a novel computational framework for studying macroscopic traveling waves in the mouse cortex by integrating realistic brain connectivity data with large-scale neural simulations.

The key contributions include:

(1) developing an algorithm that combines spatial transcriptomic data (providing detailed neuron positions and molecular properties) with voxelized connectivity data from the Allen Brain Atlas to construct neuron-to-neuron connections across ~300,000 cortical neurons;

(2) building a GPU-accelerated simulation platform capable of modeling this large-scale network with both excitatory and inhibitory Hodgkin-Huxley neurons;

(3) extending phase-based analysis methods from 2D to 3D to quantify traveling wave activity in the realistic brain geometry; and

(4) demonstrating that realistic Allen connectivity generates significantly higher levels of macroscopic traveling waves compared to simplified local or uniform connectivity patterns.

The study reveals that wave activity depends non-monotonically on coupling strength and that slow oscillations (0.5-4 Hz) are particularly conducive to large-scale wave propagation, providing new insights into how anatomical connectivity enables flexible spatiotemporal dynamics across the cortex.

Strengths:

The authors leverage two existing dense datasets of spatial transcriptomic data and connection strength between pairwise voxels in the mouse cortex in a novel way, allowing for the computational model to capture molecular and functional properties of neurons as determined by their neurotransmitter profiles, rather than making arbitrary assignments of excitatory/inhibitory roles. Additionally, the author's expansion of 2D phase dynamics to 3D phase gradient analysis methods is important and can be widely applied to calcium imaging, LFP recordings, and likely other electrophysiological recordings.

Weaknesses:

Despite these important computational advancements, a few aspects of this model, particularly the inability to validate the model with experimental neural data, diminish my enthusiasm for this paper:

(1) The model's Allen connectivity approach overlooks critical aspects of real cortical dynamics. Most importantly, it excludes subcortical structures, especially the thalamus, which drives cortical traveling waves through thalamocortical interactions. The authors' method of electrically stimulating all layer 4 neurons simultaneously to initiate waves is artificially crude and bears little resemblance to natural wave generation mechanisms.

(2) The model handles voxel-to-voxel connections crudely when neurons have mixed excitatory/inhibitory properties and varying synaptic strengths. Real connectivity differs dramatically between neuron types (pyramidal cells vs. interneurons, across cortical layers), but the model only distinguishes excitatory and inhibitory neurons. Additionally, uniform synaptic weights ignore natural variations in connection strength based on neuron type, distance, and functional role. Integrating the updated thalamocortical dataset mentioned by the authors, even at regional resolution, would substantially improve the model.

(3) While the authors bridge microscopic (single neuron) and mesoscopic (regional connectivity) data to study macroscopic (whole-cortex) waves, they don't integrate the distinct mechanisms operating at each scale. The framework demonstrates that realistic connectivity enables macroscopic waves but fails to connect how wave dynamics emerge and interact across spatial scales systematically.

(4) Claims that Allen connectivity produces higher phase gradient directionality (PGD) than local connectivity appear limited to delta oscillations at very specific coupling strengths and applied currents. Few parameter combinations show significantly higher PGD for Allen connectivity, and these are generally low PGD values overall.

(5) Broadly, it's unclear how this computational framework can study memory, learning, sleep, sensory processing, or disease states, given the disconnect between simulated intracellular voltages and the local field potentials or other electrophysiological measurements typically used to study cortical traveling waves. While computationally impressive, the practical research applications remain vague.

(6) The paper needs a clearer explanation for why medium coupling (100%) eliminates waves in Allen connectivity (Figure 6) while stronger coupling (150%) restores them.

(7) Does using a single connectivity parameter (ρ = 300) across all regions miss important regional differences in cortical connectivity density?

Reviewer #2 (Public review):

Summary:

This work presents a spiking network model of traveling waves at the whole-brain scale in the mouse neocortex. The authors use data from the Allen Institute to reconstruct connectivity between different neocortical sites. They then quantify macroscopic traveling waves following stimulation of all layer 4 neurons in the neocortex.

Strengths:

Overall, the results are interesting and shed new light on the dynamic organization of activity across the neocortex of the mouse. The paper uses realistic neuron models specifically fit to intracellular recordings, demonstrating that traveling waves occur in the mouse neocortex with both realistic connectivity and realistic single-neuron dynamics. The paper is also well-written in general. For these reasons, the authors have generally achieved their aims in this work.

Weaknesses:

(1) Description of Algorithm 1:

While the Methods section clearly explains the density parameter \rho, the statement on line 358 concerning the "ideal" average number of connections is a little unclear. The authors should explicitly clarify that \rho is a free parameter that can be adjusted to balance computational feasibility (for a given set of computational resources) and biological fidelity.

(2) Lines 102-103:

The \rho parameter used here results in approximately 300 connections per neuron on average. The authors should state clearly that the number of connections per cell is the key determinant of computational feasibility (cf. Morrison et al., Neural Computation, 2005). The authors should also review neuronal density and synaptic connectivity in the mouse neocortex and clearly reference density and connectivity in their model to the biological scales found in the mouse.

(3) Line 131:

From the plots in Figure 2, it is not clear that the stimulus response is necessarily a rhythmic oscillation, in the sense of a single narrowband frequency.

(4) Line 217:

The authors should clarify how these findings relate to the results from Mohajerani et al. (Nature Neuroscience, 2013) or differ from them.

(5) Line 230:

Because higher temporal frequency activity also tends to be more spatially localized, a correlation between PGD and temporal frequency could be an inherent consequence of this relationship, rather than a meaningful result.

(6) Line 247-248:

It is not clear that the algorithm for generating connections between neurons presented here really relates to those for community detection. For example, in the case of the Allen Institute data, the communities are essentially in the data already.

(7) Line 284-285:

The relationship between conduction delay is more direct than this sentence suggests. Conduction delay is fundamentally determined by the time required for action potentials to propagate along axons, making it intrinsically linked to anatomical distance.

(8) Line 287-288:

The authors suggest at this point that they do not have enough information to estimate time delays due to axonal conduction along white matter fibers. However, experimental data from white matter connections typically includes information about fiber length, which does enable estimating conduction delays. These estimations have been previously implemented for Allen Institute connectome data in the mouse (Choi and Mihalas, PLoS Comput Biology, 2019) and human connectome data (Budzinski et al., Physical Review Research, 2023).

(9) Lines 294-295:

Several methods do exist for detecting and characterizing wave dynamics in three-dimensional data (Budzinski et al., Physical Review Research, 2023).

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

(1.1) The manuscript “Realistic coupling enables flexible macroscopic traveling waves in the mouse cortex” by Sun, Forger, and colleagues presents a novel computational framework for studying macroscopic traveling waves in the mouse cortex by integrating realistic brain connectivity data with large-scale neural simulations.

The key contributions include: (1) developing an algorithm that combines spatial transcriptomic data (providing detailed neuron positions and molecular properties) with voxelized connectivity data from the Allen Brain Atlas to construct neuron-to-neuron connections across 300,000 cortical neurons; (2) building a GPU-accelerated simulation platform capable of modeling this large-scale network with both excitatory and inhibitory HodgkinHuxley neurons; (3) extending phase-based analysis methods from 2D to 3D to quantify traveling wave activity in the realistic brain geometry; and (4) demonstrating that realistic Allen connectivity generates significantly higher levels of macroscopic traveling waves compared to simplified local or uniform connectivity patterns.

The study reveals that wave activity depends non-monotonically on coupling strength and that slow oscillations (0.5-4 Hz) are particularly conducive to large-scale wave propagation, providing new insights into how anatomical connectivity enables flexible spatiotemporal dynamics across the cortex.

The authors leverage two existing dense datasets of spatial transcriptomic data and connection strength between pairwise voxels in the mouse cortex in a novel way, allowing for the computational model to capture molecular and functional properties of neurons as determined by their neurotransmitter profiles, rather than making arbitrary assignments of excitatory/inhibitory roles. Additionally, the author’s expansion of 2D phase dynamics to 3D phase gradient analysis methods is important and can be widely applied to calcium imaging, LFP recordings, and likely other electrophysiological recordings.

Thank you for the accurate summary of our manuscript and list of strengths.

(1.2) The model’s Allen connectivity approach overlooks critical aspects of real cortical dynamics. Most importantly, it excludes subcortical structures, especially the thalamus, which drives cortical traveling waves through thalamocortical interactions. The authors’ method of electrically stimulating all layer 4 neurons simultaneously to initiate waves is artificially crude and bears little resemblance to natural wave generation mechanisms.

We agree that excluding subcortical structures, especially the thalamus, is an important limitation of the current model. Because adding these structures would substantially expand the scope and complexity of the model, we now state this limitation more explicitly in the Discussion and leave it as a future extension:

“For simulations, we choose to randomly stimulate the total population of layer 4 neurons as a way to mimic subcortical input and generate traveling waves, which can be unrealistic. Subcortical structures, such as the thalamus, are vital to cortical dynamics like slow-wave activity [1] and are known to regulate traveling waves [2]. Therefore, a direct and important future improvement would be adding subcortical structures to the model.”

We also agree that the original constant-current stimulation was too artificial. We therefore replaced it with a 10Hz Poisson spike train delivered to layer-4 excitatory neurons across the isocortex, which more closely mimics the stochastic input that cortex receives from subcortical regions such as the thalamus. The revised stimulation protocol is described in the Results:

“Therefore, to mimic the stochastic drive that cortex receives from subcortical regions like thalamus, we deliver a 10Hz Poisson spike train to layer-4 excitatory neurons (Figure 2a), since layer 4 is the canonical thalamocortical input layer [3]; each Poisson event applies a fixed voltage bump Vstim, and no other external input is applied.”

as well as in the Methods 4.5 Simulation.

The revised stimulation protocol allowed us to rerun the parameter sweep under stochastic drive. A direct comparison of alternative wave-generating mechanisms remains an important direction for future work.

(1.3) The model handles voxel-to-voxel connections crudely when neurons have mixed excitatory/inhibitory properties and varying synaptic strengths. Real connectivity differs dramatically between neuron types (pyramidal cells vs. interneurons, across cortical layers), but the model only distinguishes excitatory and inhibitory neurons. Additionally, uniform synaptic weights ignore natural variations in connection strength based on neuron type, distance, and functional role. Integrating the updated thalamocortical dataset mentioned by the authors, even at regional resolution, would substantially improve the model.

We thank the reviewer for raising these important points regarding cell-type-specific connectivity and heterogeneity of synaptic weights. We agree that cell-type-specific connectivity and heterogeneous synaptic weights are important limitations. Because the current voxelized projectome is not cell-type specific, we now state these limitations explicitly and outline how future versions of the model could incorporate improved density and synaptic weight assumptions in the Discussion (Construction of the Allen connectivity). Specifically, we now write:

“First, the voxelized projection data is not cell-type specific. In our model, we distinguish only two neuronal populations: glutamatergic (excitatory) and GABAergic (inhibitory), based on the Zhuang-ABCA-1 transcriptomic dataset. However, real cortical connectivity differs dramatically between more refined cell types: pyramidal neurons and interneurons have distinct connection targets, and connectivity is strongly layer-specific. At the same time, we use a single global density parameter ρ to set the average number of connections per neuron across the entire cortex. A region-specific ρ could better capture the known variation in local synaptic density across cortical areas, for example the higher synaptic density of primary sensory regions relative to higher-order association areas. Future versions of the model could allow a region- and cell-type-specific ρ derived from region-level synapse-density atlases together with cell-type-resolved connectivity [4], which would simultaneously address the limitations noted above.

Also, our model uses uniform synaptic weights within each synapse type: every AMPA synapse has conductance gAMPA and every GABA synapse has conductance gGABA. In reality, synaptic strength varies with presynaptic and postsynaptic cell type, anatomical distance, and the distribution of synaptic weights is typically heavy-tailed (e.g. log-normal [5]). Extending our algorithm to sample synaptic weights from realistic distributions would be a natural next step and is likely necessary for quantitative comparisons with electrophysiological recordings.”

These additions clarify which aspects of the current model are constrained by the available voxelized projectome and which extensions would require cell-type-resolved connectivity data, region-specific density information, or more detailed synaptic-weight estimates.

(1.4) While the authors bridge microscopic (single neuron) and mesoscopic (regional connectivity) data to study macroscopic (whole-cortex) waves, they don’t integrate the distinct mechanisms operating at each scale. The framework demonstrates that realistic connectivity enables macroscopic waves but fails to connect how wave dynamics emerge and interact across spatial scales systematically.

We thank the reviewer for this insightful comment. In the revision, we added the Kuramoto synchrony analysis as a first step toward connecting scales: changes in the microscopic coupling parameter alter network synchrony, and this synchrony measure is closely associated with the macroscopic PGD observable (new Fig. 5). We agree that a full separation of layer-specific microcircuit mechanisms, region-specific connectivity motifs, and whole-cortex wave propagation remains beyond the scope of the present study. The revised text now frames the synchrony-to-wave relationship as one concrete cross-scale link that can be investigated further with this framework.

(1.5) Claims that Allen connectivity produces higher phase gradient directionality (PGD) than local connectivity appear limited to delta oscillations at very specific coupling strengths and applied currents. Few parameter combinations show significantly higher PGD for Allen connectivity, and these are generally low PGD values overall.

We agree that the original comparison did not sufficiently establish whether the Allen-connectivity advantage held beyond a small number of parameter choices.

In the revised manuscript, we expanded the analysis to a 10 × 12 grid of excitatory coupling strengths and Poisson stimulus magnitudes for Allen, local, and uniform connectivity. Across this full (gAMPA,Vstim) grid, Allen connectivity shows higher per-band maximum PGD than local or uniform connectivity, especially in the theta, alpha, and beta bands (Fig. 5b). The delta-band difference is small, consistent with the reviewer’s observation that the original delta-band result did not clearly separate Allen from local connectivity.

In the revised manuscript, Fig. 5c and Fig. 5d show how PGD varies with stimulus magnitude and coupling strength individually. The full per-band PGD heatmaps for all three connectivities, together with the per-band Allen-minus-opponent gap bars, are provided in Figure 4-figure supplement 6. These additions show that the Allen-connectivity trend is not limited to a single representative operating point.

(1.6) Broadly, it’s unclear how this computational framework can study memory, learning, sleep, sensory processing, or disease states, given the disconnect between simulated intracellular voltages and the local field potentials or other electrophysiological measurements typically used to study cortical traveling waves. While computationally impressive, the practical research applications remain vague.

We thank the reviewer for this important point. To bridge the gap between our simulations and experimentally measured data, such as local field potentials (LFP), we used a post-hoc LFP estimation pipeline and added a dedicated Methods subsection describing it (LFP estimation from intracellular voltage). The key idea is that LFP primarily reflects the net transmembrane synaptic current in a local population, which we can reconstruct directly from the voltage traces and the connectivity used in the simulation. Please see Methods section LFP estimation from intracellular voltage.

This pipeline allows us to test whether the macroscopic traveling-wave structure identified in the voltage traces is also present in an LFP-like signal. In the revised manuscript, we added a new Results section and show, in Fig. 6, side-by-side voltage- and LFP-based snapshots, the voltage-LFP PGD scatter (Pearson r = 0.78-0.89 per band), and per-band PGD comparisons across Allen, local, and uniform connectivity on the LFP signal. These results indicate that our conclusions are not restricted to intracellular voltage and provide a closer bridge to LFP-based experimental measurements.

(1.7) The paper needs a clearer explanation for why medium coupling (100%) eliminates waves in Allen connectivity (Figure 6) while stronger coupling (150%) restores them.

Thank you for requesting this clarification. The revised analysis suggests that the non-monotonic relationship between coupling strength and wave activity reflects an interaction between network synchrony and spatial organization. At weak coupling, the network has enough coordination to support propagating waves. At medium coupling, increased synaptic drive pushes the network into an asynchronous irregular state that disrupts coherent wave fronts. At strong coupling, rhythmic synchronization is re-established and again supports wave propagation.

To substantiate this explanation quantitatively, we measured the Kuramoto order parameter R(t)=|〈eiϕ(x,t)〉x| from the generalized phase ϕ(x, t) of the band passed voltage field (see updated Quantitative measurement of neuronal activity in Methods) and reduced it to the maximum over each 1-s recording window. We then swept the same ten gAMPA values (0.005-0.050 nS) and twelve stimulus magnitudes used for the PGD sweeps, for Allen, local and uniform connectivity, in all five frequency bands. The new analysis is presented in Fig. 5: panel (e) shows synchrony versus gAMPA for the three connectivities, panel (d) shows the matching PGD curve, and panel (f) shows the synchrony-PGD scatter with Pearson r per connectivity.

The synchrony curve captures the main peak-dip-recovery structure of the PGD curve. For Allen connectivity in the alpha band, mean synchrony peaks at weak coupling, collapses to a 75 % suppression in the medium-coupling window gAMPA = 0.020-0.035 nS, and recovers near unity at gAMPA ≥ 0.040 nS. Uniform connectivity follows the same U-shape with a slightly earlier dip. Per-band versions of the PGD- and synchrony-vs-coupling curves are provided in Figure 5-figure supplement 1, showing that the peak-dip-recovery profile holds across all five canonical bands for Allen and uniform connectivity. Across the 120- point (gAMPA, Vstim) grid, synchrony and PGD are positively correlated in every band and every connectivity (Pearson 0r ranging from ≈ 0.30 to ≈ 0.92 across band-connectivity combinations; per-band scatters in Figure 5-figure supplement 2). Local connectivity follows a different trajectory: its synchrony is moderate at weak coupling and decays monotonically with gAMPA without recovering at strong coupling. This is consistent with local connectivity not supporting large-scale propagating waves at strong coupling, so the peak-dip-recovery interpretation applies mainly to Allen and uniform connectivity. We have added this synchrony analysis to the revised manuscript:

“To diagnose the mechanism behind this profile, we measured the Kuramoto order parameter R(t) = |⟨eiϕ(x,t)⟩x| from the generalized phase field ϕ(x, t) of each frequency band and recorded its maximum over each simulation window (Quantitative measurement of neuronal activity). The resulting synchrony curve (Figure 5e) resembles the trend of the maximum PGD well (Figure 5d). For Allen connectivity in the alpha band, mean synchrony peaks at weak coupling, collapses in the medium-coupling window, and recovers at strong coupling. Across the full 120-point (gAMPA, Vstim) grid, synchrony and PGD are positively correlated in every band and every connectivity (Pearson r = 0.30- 0.92, all p < 10−3; Figure 5f, with per-band scatters in figure Supplement 2). Therefore, the PGD trough at medium coupling may be a synchrony trough: increased synaptic drive pushes the network into an asynchronous irregular state, while strong coupling re-establishes rhythmic synchronization that supports wave propagation. This synchrony-to-wave bottleneck seems more significant to the networks with long-range connectivity (Allen, uniform), partly because long-range connections can augment synchrony in the coupled neuronal network [6]. We also note that although uniform connectivity is able to achieve almost an identical level of synchrony to that of Allen connectivity, the PGD remains much lower due to the loss of spatial organization within.”

(1.8) Does using a single connectivity parameter (ρ = 300) across all regions miss important regional differences in cortical connectivity density?

We thank the reviewer for raising this important point. We agree that a single global density parameter misses region-to-region variation in local synaptic density, and we have extended the Discussion (Construction of the Allen connectivity) to state this limitation alongside the cell-type-specific connectivity and synaptic-weight limitations discussed in response to Point 1.3. Specifically, we now write:

“At the same time, we use a single global density parameter ρ to set the average number of connections per neuron across the entire cortex. A region-specific ρ could better capture the known variation in local synaptic density across cortical areas, for example the higher synaptic density of primary sensory regions relative to higher-order association areas. Future versions of the model could allow a region- and cell-type-specific ρ derived from region-level synapse density atlases together with cell-type-resolved connectivity [4], which would simultaneously address the limitations noted above.”

This paragraph is placed directly after the cell-type and weight-heterogeneity limitations added in response to Point 1.3.

Reviewer #2 (Public review):

(2.1) This work presents a spiking network model of traveling waves at the whole-brain scale in the mouse neocortex. The authors use data from the Allen Institute to reconstruct connectivity between different neocortical sites. They then quantify macroscopic traveling waves following stimulation of all layer 4 neurons in the neocortex.

Overall, the results are interesting and shed new light on the dynamic organization of activity across the neocortex of the mouse. The paper uses realistic neuron models specifically fit to intracellular recordings, demonstrating that traveling waves occur in the mouse neocortex with both realistic connectivity and realistic single-neuron dynamics. The paper is also well-written in general. For these reasons, the authors have generally achieved their aims in this work.

We thank Reviewer 2 for the positive assessment and accurate summary of our work.

(2.2) Description of Algorithm 1: While the Methods section clearly explains the density parameter ρ, the statement on line 358 concerning the “ideal” average number of connections is a little unclear. The authors should explicitly clarify that ρ is a free parameter that can be adjusted to balance computational feasibility (for a given set of computational resources) and biological fidelity. The ρ parameter used here results in approximately 300 connections per neuron on average. The authors should state clearly that the number of connections per cell is the key determinant of computational feasibility (cf. Morrison et al., Neural Computation, 2005). The authors should also review neuronal density and synaptic connectivity in the mouse neocortex and clearly reference density and connectivity in their model to the biological scales found in the mouse.

We thank the reviewer for raising this important point about our connectivity algorithm and simulation. We have clarified in the revised Methods (Use of the voxelized connectivity data, Methods 4.2) that ρ is a free parameter that controls the trade-off between computational feasibility and biological fidelity. Specifically, we now write:

“The density parameter ρ is a free parameter that controls the trade-off between computational feasibility and biological fidelity: higher values of ρ yield more connections per neuron and thus higher biological realism, at the cost of greater memory and runtime [7].”

A careful accounting of the biological scales involved (synapse density per neuron, total cortical population) and incorporating region or cell-type-specific connection density is left as a future direction. We have noted this in the Discussion subsection:

“First, the voxelized projection data is not cell-type specific. In our model, we distinguish only two neuronal populations: glutamatergic (excitatory) and GABAergic (inhibitory), based on the Zhuang-ABCA-1 transcriptomic dataset. However, real cortical connectivity differs dramatically between more refined cell types: pyramidal neurons and interneurons have distinct connection targets, and connectivity is strongly layer-specific. At the same time, we use a single global density parameter ρ to set the average number of connections per neuron across the entire cortex. A region-specific ρ could better capture the known variation in local synaptic density across cortical areas, for example the higher synaptic density of primary sensory regions relative to higher-order association areas. Future versions of the model could allow a region- and cell-type-specific ρ derived from region-level synapse-density atlases together with cell-type-resolved connectivity [4], which would simultaneously address the limitations noted above.”

(2.3) Line 131: From the plots in Figure 2, it is not clear that the stimulus response is necessarily a rhythmic oscillation, in the sense of a single narrowband frequency.

The reviewer is correct, and we are grateful for the prompt to be more precise. The Results phrasing around Fig. 2 has been softened to avoid any implication of narrowband rhythmicity, and now reads:

“Under this protocol, we immediately observe macroscopic traveling waves emerge across the cortex (Figure 2 and Videos). The global mean voltage and the region-sorted raster (Figure 2c, d) reveal oscillatory activity that is well synchronized across regions, while the local-mean intracellular voltage maps over a representative 50ms window (Figure 2b) reveal a coherent wavefront sweeping along the anterior-posterior axis, consistent with previously reported cortex-wide waves [8, 9, 10]. The corresponding single-neuron-resolution view of the same simulation, with no spatial averaging, is shown in figure Supplement 1.”

Moreover, we have revised the Introduction to describe [11] as demonstrating traveling waves in broadband (5-40Hz) activity, making clear that traveling waves can occur without requiring narrowband oscillations (see also our response to your related point below). In the revised manuscript, we also separate the broadband activity into canonical frequency bands and analyze the wave activity in each band independently.

(2.4) Line 217: The authors should clarify how these findings relate to the results from Mohajerani et al. (Nature Neuroscience, 2013) or differ from them.

We thank the reviewer for this suggestion. We agree that [12] is an important experimental benchmark. A direct quantitative comparison is difficult because the original data were not aligned to the Allen Brain Atlas CCF used in our simulations. We therefore revised the Discussion to identify this comparison as a future direction, alongside the visual-cortex bidirectional-wave data of [13]:

“A more detailed quantitative comparison with experimental cortical-wave studies, such as the cortex-wide voltage-imaging data of [12] or the bidirectional visual-evoked waves reported by [13], is left as a future direction.”

(2.5) Line 230: Because higher temporal frequency activity also tends to be more spatially localized, a correlation between PGD and temporal frequency could be an inherent consequence of this relationship, rather than a meaningful result.

We thank the reviewer for raising this important point. The reviewer is correct that higher-frequency oscillations tend to be more spatially localized, which can inherently reduce PGD when measured globally. We therefore revised this analysis by separating the broadband signals into canonical frequency bands and comparing PGD within each band.

In the revised manuscript, we no longer interpret cross-band PGD differences as evidence for a frequency-to-spatial-scale relationship. Instead, we report the level of macroscopic wave activity within each canonical band. This per-band comparison is summarized in Fig. 5b, which reports the maximum PGD in each band (mean ± SEM across the entire (gAMPA,Vstim) grid) for the three connectivities. Allen connectivity shows higher per-band PGD than local or uniform connectivity in the theta, alpha, and beta bands, without requiring an interpretation of PGD differences across frequency bands.

For completeness, we also computed the per-band mean(Allen) − mean(opponent) PGD gap across the full parameter grid (10 gAMPA × 12 Vstim = 120 points per connectivity). The result is presented in Figure 4-figure supplement 6: panel (a) gives the per-band maxPGD heatmaps over the (gAMPA, Vstim) grid for each connectivity, and panel (b) gives the per-band Allen-minus-opponent mean-PGD gap. The mean gap against Local is −0.004 in delta, +0.027 in theta, +0.036 in alpha, +0.036 in beta and +0.004 in gamma, and against Uniform is +0.015, +0.044, +0.052, +0.040 and +0.011, respectively. The gap is largest in alpha and broadly concentrated in theta-alpha-beta, rather than in delta as we had originally written. We report these per-band gaps descriptively because they summarize one parameter sweep per connectivity rather than independent biological or simulation replications, and we do not interpret the differences across bands as a meaningful frequency dependence. We have added this analysis to the revised manuscript.

(2.6)Line 247-248: It is not clear that the algorithm for generating connections between neurons presented here really relates to those for community detection. For example, in the case of the Allen Institute data, the communities are essentially in the data already.

We agree with the reviewer that the relevant anatomical blocks are already present in the data. Our intent was to relate the sampling procedure to stochastic block models for network generation, not to community-detection algorithms. We have revised this passage in the Discussion subsection Construction of the Allen connectivity: ”In essence, our algorithm belongs to the family of stochastic block models [14], where the block structure is given by the voxelization of the Allen Brain Atlas and the inter-block connection probabilities are set by the voxelized projection strengths.”

(2.7) Line 284-285: The relationship between conduction delay is more direct than this sentence suggests. Conduction delay is fundamentally determined by the time required for action potentials to propagate along axons, making it intrinsically linked to anatomical distance.

Thank you for raising this important point. We agree that conduction delay is directly tied to axonal propagation time and therefore to anatomical path length. Our original wording was intended to note that the relevant path length cannot be approximated reliably by Euclidean distance in the 3-D coordinate space. We have revised this passage in the Discussion sub-section Cortical model and simulation to state that conduction delay is linked to the white-matter path length of the connection, while noting that our model lacks the actual axonal path geometry through the cortical manifold:

“However, we did not include conduction delay in our study. Conduction delay is thought to have a proportional relationship with the white matter path length of the connection between two neurons [15]. In our model, although we have the 3D positions of neurons, we do not have geometric information about the cortical manifold. Two neurons can be very close in the Cartesian coordinates measured by the Euclidean distance, but very far in terms of the length of the actual connection in the brain. Therefore, incorporating accurate conduction delay in the model is an important future direction.”

(2.8) Lines 294-295: Several methods do exist for detecting and characterizing wave dynamics in three-dimensional data (Budzinski et al., Physical Review Research, 2023).

Thank you for this reference. We have added a citation to [16] in the Discussion subsection Quantitative measurements of 3-D traveling waves, acknowledging that methods for 3-D wave analysis do exist while noting that most published algorithms are designed for 2-D data:

“There are many techniques available for identifying and measuring large-scale neuronal spatiotemporal patterns [17, 18]. While methods for detecting wave dynamics in three-dimensional data do exist [16], most published algorithms are designed to analyze 2-D data, so our simulation data, which is intrinsically 3-D, presents new challenges for measurement.”

(2.9) Line 28: It is important to note that the Davis et al. (2020) reference is not actually in the beta band, but instead in the broadband (5-40 Hz). This distinction is important because it demonstrates that waves can occur in neural data without requiring narrowband oscillations.

Thank you for this correction. We have moved the [11] citation out of the beta-band group in the Introduction and reframed it as a broadband (5-40Hz) reference, so that the sentence now reads:

“These waves are observed at different frequencies during various brain activities, ranging from slow-wave activity [8, 9], sleep spindles [19], to faster oscillations in alpha [20, 21], beta [22, 23], and gamma [21, 13] frequency bands, as well as in broadband (5-40Hz) activity [11].”

This distinction is important because it makes clear that traveling waves do not require narrowband oscillations. The revised manuscript therefore separates the broadband activity into canonical frequency bands and compares wave activity within each band.

(2.10) Line 46-49: This sentence could be clearer, for example, by specifying “certain dynamics” in more precise terms.

We apologize for the imprecision in the original manuscript. We have clarified the sentence in the Introduction to specify that the dynamics of interest are coexistence patterns of local and global activity in the network, as described in the cited reference [24]:

“It has been shown that in a coupled neuronal network, the coexistence of global wave activity with locally asynchronous states only arises when the number of oscillators is high enough, where local and global activities can coexist [24].”

(2.11) Figure 1b(i): Small typo in the label for this panel.

This has been corrected. Thank you for catching it.

(2.12) Lines 121-122: It may be important to note that spiking neural networks can also generate self-sustained activity (Vogels and Abbott, JNeurosci, 2005; Kumar et al., Neural Computation, 2008). This self-sustained activity is a form of internally generated “frozen” noise that is fundamentally different from externally imposed noise sources (such as Poisson external input) (Destexhe and Contreras, Science, 2006). Waves appear in this self-sustained activity, as well (Davis et al., Nature Communications, 2021), supporting the generality of this phenomenon.

Thank you for pointing out these important references. We have added citations to [25], [26], [27], and [24] in the Results subsection Macroscopic traveling waves emerge from random stimulation through realistic connectivity:

“Spiking networks of this scale can also generate self-sustained activity in similar regimes [25, 26], which differs fundamentally from externally imposed noise [27], and traveling waves have been reported under such conditions [24].”

In the revised manuscript, we also changed the stimulation protocol from constant current to Poisson input, which is more similar to the stochastic input that cortical neurons receive in vivo. Traveling waves observed in our model under this Poisson drive therefore complement, rather than depend on, the self-sustained-activity regime emphasized by the cited works.

(2.13) Lines 139-140: “anterior-posterior macroscopic waves in both directions” and ”in the reverse direction right after each other” could be clearer. In addition, the study from Aggarwal et al. (Nature Communications, 2022) could be relevant to note at this point.

We thank the reviewer for these wording suggestions and the relevant reference. In the revised manuscript, we reran the simulations and updated Figure 2 accordingly. The new representative simulation shown in Fig. 2 emphasizes a coherent wavefront sweeping along the anterior-to-posterior axis, and the original passages describing consecutive opposite-direction waves have been removed from the Results subsection Macroscopic traveling waves emerge from random stimulation through realistic connectivity, which now reads:

“Under this protocol, we immediately observe macroscopic traveling waves emerge across the cortex (Figure 2 and Videos). The global mean voltage and the region-sorted raster (Figure 2c, d) reveal oscillatory activity that is well synchronized across regions, while the local-mean intracellular voltage maps over a representative 50ms window (Figure 2b) reveal a coherent wavefront sweeping along the anterior-posterior axis, consistent with previously reported cortex-wide waves [8, 9, 10]. The corresponding single-neuron-resolution view of the same simulation, with no spatial averaging, is shown in figure Supplement 1.”

Bidirectional propagation can still occur in the model, but we have chosen not to present it as a focal result of this revised manuscript; consequently, the specific phrasings flagged by the reviewer no longer appear in the Results text.

We agree that [13] is an important reference, and we now discuss it in the Comparing with experimental data subsection as a potential benchmark for future quantitative comparison with our model.

(2.14) Figure 7: The caption for this figure could be clearer.

Line 281: typo ”Ermentrou”.

Line 418: typo ”excitatory”.

The typos (“Ermentrou” → “Ermentrout”, and the “excitatory” typo at line 418) have been corrected. The original Figure 7 has been removed from the revised manuscript; its content (PGD versus stimulus magnitude and coupling strength, and PGD by dominant frequency bucket) has been reorganized across the new Figs. 4-6.

References

(1) Steriade M, Mccormick DA, Sejnowski TJ. Thalamocortical Oscillations in the Sleeping and Aroused Brain. Science. 1993;262(5134):679-85. Available from: :// WOS:A1993MD95200029.

(2) Ye Z, Bull MS, Li A, Birman D, Daigle TL, Tasic B, et al. Brain-wide topographic coordination of traveling spiral waves. bioRxiv. 2023:2023-12.

(3) Rockland KS.What do we know about laminar connectivity? Neuroimage. 2019;197:772-84.

(4) Harris JA, Mihalas S, Hirokawa KE, Whitesell JD, Choi H, Bernard A, et al. Hierarchical organization of cortical and thalamic connectivity. Nature. 2019;575(7781):195+. Available from: ://WOS:000496159900061https://www.nature.com/articles/ s41586-019-1716-z.pdf.

(5) Buzs´aki G, Mizuseki K. The log-dynamic brain: how skewed distributions affect network operations. Nature Reviews Neuroscience. 2014;15(4):264-78.

(6) Bazhenov M, Rulkov NF, Timofeev I. Effect of synaptic connectivity on long-range synchronization of fast cortical oscillations. Journal of neurophysiology. 2008;100(3):156275.

(7) Morrison A, Aertsen A, Diesmann M. Spike-timing-dependent plasticity in balanced random networks. Neural computation. 2007;19(6):1437-67.

(8) Massimini M. The Sleep Slow Oscillation as a Traveling Wave. Journal of Neuroscience. 2004;24(31):6862-70. Available from: https://dx.doi.org/10.1523/jneurosci. 1318-04.2004 https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6729597/pdf/ 0246862.pdf.

(9) Liang Y, Song C, Liu M, Gong P, Zhou C, Kno¨pfel T.Cortex-Wide Dynamics ofIntrinsic Electrical Activities: Propagating Waves and Their Interactions. The Journal of Neuroscience. 2021;41(16):3665-78. Available from: https://www.jneurosci.org/ content/jneuro/41/16/3665.full.pdf.

(10) Aggarwal A, Luo J, Chung H, Contreras D, Kelz MB, Proekt A. Neural assemblies coordinated by cortical waves are associated with waking and hallucinatory brain states. Cell Reports. 2024;43(4):114017. Available from: https://www.sciencedirect.com/ science/article/pii/S2211124724003450.

(11) Davis ZW, Muller L, Martinez-Trujillo J, Sejnowski T, Reynolds JH. Spontaneous travelling cortical waves gate perception in behaving primates. Nature. 2020;587(7834):4326. Available from: https://doi.org/10.1038/s41586-020-2802-y.

(12) Mohajerani MH, Chan AW, Mohsenvand M, LeDue J, Liu R, McVea DA, et al. Spontaneous cortical activity alternates between motifs defined by regional axonal projections. Nature Neuroscience. 2013;16(10):1426-35. Available from: https://doi.org/ 10.1038/nn.3499https://www.nature.com/articles/nn.3499.pdf.

(13) Aggarwal A, Brennan C, Luo J, Chung H, Contreras D, Kelz MB, et al. Visual evoked feedforward–feedback traveling waves organize neural activity across the cortical hierarchy in mice. Nature Communications. 2022;13(1):4754. Available from: https://doi.org/10.1038/s41467-022-32378-xhttps://www.nature. com/articles/s41467-022-32378-x.pdf.

(14) Holland PW, Laskey KB, Leinhardt S. Stochastic blockmodels: First steps. Social Networks. 1983;5(2):109-37. Available from: https://www.sciencedirect.com/science/ article/pii/0378873383900217.

(15) Lemar´echal JD, Jedynak M, Trebaul L, Boyer A, Tadel F, Bhattacharjee M, et al. A brain atlas of axonal and synaptic delays based on modelling of cortico-cortical evoked potentials. Brain. 2022;145(5):1653-67.

(16) Budzinski RC, Nguyen TT, Min´o-Calero J, Davis ZW, Muller LE. Analyzing transientevoked neural activity in three-dimensional cortical recordings. Physical Review Research. 2023;5(1):013012.

(17) Townsend RG, Gong P. Detection and analysis of spatiotemporal patterns in brain activity. PLOS Computational Biology. 2018;14(12):e1006643. Available from: https: //doi.org/10.1371/journal.pcbi.1006643.

(18) Gutzen R, De Bonis G, De Luca C, Pastorelli E, Capone C, Allegra Mascaro AL, et al. A modular and adaptable analysis pipeline to compare slow cerebral rhythms across heterogeneous datasets. Cell Reports Methods. 2024;4(1). Available from: https: //doi.org/10.1016/j.crmeth.2023.100681.

(19) Muller L, Piantoni G, Koller D, Cash SS, Halgren E, Sejnowski TJ. Rotating waves during human sleep spindles organize global patterns of activity that repeat precisely through the night. Elife. 2016;5. Available from: https://www.ncbi.nlm.nih.gov/pubmed/27855061https://www.ncbi.nlm. nih.gov/pmc/articles/PMC5114016/pdf/elife-17267.pdf.

(20) Zhang H, Watrous AJ, Patel A, Jacobs J. Theta and alpha oscillations are traveling waves in the human neocortex. Neuron. 2018;98(6):1269-81. e4.

(21) van Kerkoerle T, Self MW, Dagnino B, Gariel-Mathis MA, Poort J, van der Togt C, et al. Alpha and gamma oscillations characterize feedback and feedforward processing in monkey visual cortex. Proc Natl Acad Sci U S A. 2014;111(40):14332-41. Available from: https://www.ncbi.nlm.nih.gov/pubmed/25205811.

(22) Bhattacharya S, Brincat SL, Lundqvist M, Miller EK. Traveling waves in the prefrontal cortex during working memory. PLoS Comput Biol. 2022;18(1):e1009827.

(23) Rubino D, Robbins KA, Hatsopoulos NG. Propagating waves mediate information transfer in the motor cortex. Nature Neuroscience. 2006;9(12):1549-57. Available from: https://doi.org/10.1038/nn1802.

(24) Davis ZW, Benigno GB, Fletterman C, Desbordes T, Steward C, Sejnowski TJ, et al. Spontaneous travelling waves naturally emerge from horizontal fiber time delays and travel through locally asynchronous-irregular states. Nature Communications. 2021;12(1):6057.

(25) Vogels TP, Abbott LF. Signal propagation and logic gating in networks of integrate-and-fire neurons. Journal of Neuroscience. 2005;25(46):10786-95.

(26) Kumar A, Schrader S, Aertsen A, Rotter S. The high-conductance state of cortical networks. Neural Computation. 2008;20(1):1-43.

(27) Destexhe A, Contreras D. Neuronal computations with stochastic network states. Science. 2006;314(5796):85-90.

Version 2

Abstract

Traveling waves are ubiquitous in neuronal systems across different spatial scales. While microscopic and mesoscopic waves are relatively well studied, the emergence of macroscopic traveling waves remains less understood. Here, by modeling the mouse cortex using spatial transcriptomic and connectivity data, we show that realistic cortical connectivity can generate a significantly higher level of macroscopic traveling waves than artificial local and uniform connectivity across multiple oscillation frequency bands, with the strongest advantage appearing in the theta, alpha, and beta frequency bands. By probing the model in different dynamic regimes, we find that macroscopic wave activity depends on both network connectivity and excitatory coupling strength, with a non-monotonic dependence on coupling. Together, our work shows how flexible macroscopic traveling waves can emerge in the mouse cortex and offers a computational framework to further study traveling waves in the mouse brain at the single-cell level.

Introduction

Traveling waves have emerged as a central topic among different large-scale spatiotemporal dynamics observed in neuronal systems Muller et al. (2018). These waves are observed at different frequencies during various brain activities, ranging from slow-wave activity Massimini (2004); Liang et al. (2021), sleep spindles Muller et al. (2016), to faster oscillations in alpha Zhang et al. (2018); van Kerkoerle et al. (2014), beta Bhattacharya et al. (2022); Rubino et al. (2006), and gamma van Kerkoerle et al. (2014); Aggarwal et al. (2022) frequency bands, as well as in broadband (5-40 Hz) activity Davis et al. (2020). These waves also appear in various brain regions across species, including the hippocampus Lubenov and Siapas (2009); Patel et al. (2012); Zhang and Jacobs (2015), visual cortex Sato et al. (2012), prefrontal cortex Bhattacharya et al. (2022), motor cortex Rubino et al. (2006), or span across multiple brain regions Liang et al. (2023). More importantly, these waves have long been conjectured to regulate brain functions through mechanisms such as synaptic plasticity Ermentrout and Kleinfeld (2001); Muller et al. (2018). And researchers have recently found that traveling waves can modulate important brain functions such as memory, visual processing, and sleep Mohan et al. (2024); Aggarwal et al. (2022); Miyamoto et al. (2017).

The prevalence and important implications of traveling waves have prompted researchers to investigate their causes and organization. In these studies, computational modeling has been useful due to its flexibility. Researchers have developed both microscopic models Davis et al. (2020); Naze et al. (2015); Ermentrout and Kleinfeld (2001) and mesoscopic models Roberts et al. (2019); Liang et al. (2023) to explore the mechanical properties and computational implications of traveling waves. However, most of these studies focus on microscopic or mesoscopic waves that appear in a single brain region. Large-scale, macroscopic waves across multiple brain regions have been less studied. Moreover, these models are usually constructed with local or uniform connectivity and lack the realistic geometry and connectivity of the brain. It has been shown that in a coupled neuronal network, the coexistence of global wave activity with locally asynchronous states only arises when the number of oscillators is high enough, where local and global activities can coexist Davis et al. (2021). At the same time, it has been argued that modeling neuronal networks at single-cell resolution will yield more precise discoveries than the mean-field model Dura-Bernal et al. (2024); Deco et al. (2008). Therefore, it is ideal to study traveling waves in a large domain such as the cortex at single-cell resolution.

However, there are two significant challenges: First, a microscopic wiring diagram of the mouse brain at the whole-brain level is not available. But we do have mesoscopic connectivity data provided by the Allen Brain Atlas Oh et al. (2014); Knox et al. (2018); Harris et al. (2019), which will be used in this work. Second, simulating and measuring wave activity in a large 3D computational domain with realistic brain geometry is computationally expensive. Simulations of large-scale neuronal networks with realistic neuronal models are usually conducted on supercomputers Yamaura et al. (2020); Markram et al. (2015).

In this work, we overcome these two challenges in order to study the emergence and propagation of macroscopic traveling waves in the mouse cortex. First, we propose a network-generating algorithm that integrates a spatial transcriptomic dataset that provides the detailed spatial and molecular profile of neurons Zhang et al. (2023), with voxelized connectivity data Knox et al. (2018) of the mouse brain. We use this algorithm to create a neuron-to-neuron connectome for one hemisphere of the mouse cortex, which we call Allen connectivity. We then constructed a 3D computational model with both excitatory and inhibitory neurons on top of the Allen connectivity. To overcome the second difficulty, we built our own parallel computational framework that can simulate the constructed model with a desktop GPU.

We simulate this model under stochastic layer-4 input, and we quantify macroscopic wave activity with a three-dimensional phase-gradient analysis. Across a broad parameter sweep, Allen connectivity produces stronger macroscopic waves than local or uniform connectivity, especially in the theta, alpha, and beta bands. We further find that wave activity depends non-monotonically on excitatory coupling strength and is closely linked to network synchrony. Together, these results show how realistic cortical connectivity can support flexible macroscopic traveling waves and provide a computational framework for studying cortex-wide wave dynamics at single-cell resolution.

Results

Construction of the cortical connectivity

A recent spatial transcriptomic dataset (Zhuang-ABCA1) Zhang et al. (2023) provides detailed positions of neurons in the mouse brain along with their molecular properties, including the neuro-transmitter information (Fig. 1a, Use of the spatial transcriptomic data). To build a computational model using this data, we need to connect neurons in Zhuang-ABCA1 based on the realistic connectivity of the mouse brain that has been experimentally measured. Previously, Oh et al. (2014) and Harris et al. (2019) have provided a mesoscopic connectivity matrix, known as the “projectome”, which is derived from thousands of viral tracing experiments to measure the strength of connections between different brain regions mesoscopically. This projectome was later refined to a finer level using a statistical model introduced by Knox et al. (2018), where the strength of connections is provided on a voxel (100μm3) resolution. Because both neuronal (Zhuang-ABCA1) and the voxelized connectivity data are registered with the Allen Brain Atlas Common Coordinate Framework (ABA-CCF) Wang et al. (2020), we now introduce an algorithm that integrates the two datasets to construct a neuron-to-neuron connectivity.

Construction of the cortical connectivity.

(a) Visualization of all neurons in the hemispherical dataset we use to build the cortical connectivity. The hemisphere contains about 1 million neurons in total, including about 300,000 cortical neurons. (b) Algorithm of building the connectome: i) First, we identify neuronal excitatory/inhibitory information based on the spatial transcriptomic data. ii) We then identify the voxelized position for all the neurons. iii) Based on the voxelized projection data, we calculate the number of connections between every pair of voxels (see Use of the voxelized connectivity data). iv) Connections are created randomly between the neurons in each pair of voxels. v) Repeat for all voxels in regions of interest. vi) Complete the connectivity. (c) The connectivity matrix shows the number of connections between different cortical regions(in log 10 scale). A full version of the connectivity matrix with annotated regions is shown in figure Supplement 1. (d) Visualization of outgoing connections from sampled neurons in the primary visual cortex (VISp, left), primary auditory cortex (AUDp, middle), and the ventral anterior cingulate areas (ACAv, right).

Consider constructing a connectivity for N neurons that are contained in M voxels, and let Xi represent the physical position of the ith neuron. We have access to the voxelized projection matrix W, where Wab denotes the projection strength from voxel a to voxel b. Our connectivity algorithm requires only a single density parameter, ρ, which determines the average number of connections per neuron, thereby controlling the network’s sparsity (Use of the voxelized connectivity data). Using this information, we propose the following algorithm to construct the connectivity for the target neurons (Fig. 1b and Methods):

Applying this algorithm to all cortical neurons identified in Zhuang-ABCA1, we obtain a cortical connectivity that consists of 297,812 neurons, and we choose each neuron to have 300 connections on average, resulting in a network sparsity of around 0.1 % (Construction and storage of the connectivity). The cortical connectivity we constructed is highly heterogeneous, where the number of connections between different regions can vary across four orders of magnitude (Fig. 1c). At the same time, both the in-degree and out-degree distribution of the connectome follow a realistic log-normal distribution Buzsáki and Mizuseki (2014) (figure Supplement 1). Since the complete connectome is difficult to visualize, we show some connections from different cortical regions, including the primary visual, auditory cortex, and the ventral anterior cingulate area (Fig. 1d).

Macroscopic traveling waves emerge from random stimulation through realistic connectivity

Equipped with the Allen connectivity, we built a single-cell-resolution cortical model to test whether realistic anatomical wiring is sufficient to convert a generic stochastic input into a coordinated, macroscopic traveling wave. Each neuron is a single-compartment Hodgkin-Huxley unit Pospischil et al. (2008), with its excitatory (glutamatergic, AMPA) or inhibitory (GABAergic) identity inherited from the Zhuang-ABCA-1 transcriptomic labels, and synapses are conductance-based with first-order kinetics (Neuronal and synaptic model). The full hemispheric network of 297,812 neurons is simulated on a single GPU (Simulation).

Spiking networks of this scale can also generate self-sustained activity in similar regimes Vogels and Abbott (2005); Kumar et al. (2008), which differs fundamentally from externally imposed noise Destexhe and Contreras (2006), and traveling waves have been reported under such conditions Davis et al. (2021). Therefore, to mimic the stochastic drive that cortex receives from subcortical regions like thalamus, we deliver a 10 Hz Poisson spike train to layer-4 excitatory neurons (Figure 2a), since layer 4 is the canonical thalamocortical input layer Rockland (2019); each Poisson event applies a fixed voltage bump Vstim, and no other external input is applied.

Macroscopic traveling waves emerge in response to random layer-4 stimulation through Allen connectivity.

(a) Stimulation protocol: a 10 Hz random Poisson spike train delivers a fixed-amplitude voltage bump Vstim to all layer-4 excitatory neurons; no other external input is applied. (b) Two rows of five spatial snapshots of the local-mean (KNN = 300) intracellular voltage, sampled every 5ms across the window highlighted in (c). A coherent wavefront propagates along the anterior-to-posterior axis. (c) Global mean intracellular voltage of all neurons (dark blue) overlaid on per-region mean voltages (light traces). The red dashed lines mark the snapshot window visualized in (b). (d) Region-sorted raster plot of ∼40,000 neurons sampled evenly from major cortical regions, anterior-to-posterior. See Videos for the movie of the simulation.

Under this protocol, we immediately observe macroscopic traveling waves emerge across the cortex (Figure 2 and Videos). The global mean voltage and the region-sorted raster (Figure 2c, d) reveal oscillatory activity that is well synchronized across regions, while the local-mean intracellular voltage maps over a representative 50 ms window (Figure 2b) reveal a coherent wavefront sweeping along the anterior-posterior axis, consistent with previously reported cortex-wide waves Massi-mini (2004); Liang et al. (2021); Aggarwal et al. (2024). The corresponding single-neuron-resolution view of the same simulation, with no spatial averaging, is shown in figure Supplement 1.

Realistic connectivity produces stronger macroscopic waves than local or uniform connectivity

In past modeling studies of traveling waves, it is common to adopt uniform-random, local, or a combination of the two kinds of connectivity. Therefore, we now simulate three models differing in connectivity in order to compare and investigate whether such macroscopic waves are presented in Figure 2 only emerge from the Allen connectivity. For comparison, we constructed two additional types of connectivity that are common in past studies of traveling waves:

Uniform connectivity: Neurons are connected across the entire cortex randomly (Fig. 3b).

Local connectivity: Neurons are connected to a fixed number of their closest neighbors (Fig. 3c).

Top: schematics of Allen (a), local (b), and uniform connectivity (c); bottom: Visualization of outgoing connections from the same 50 randomly selected neurons in the primary visual cortex (VISp) for the three connectivity.

Here we visualize the outgoing connections from 50 randomly selected neurons in the primary visual cortex to show the difference between the three connectivity. Under Allen connectivity, the outgoing connections project to different regions heterogeneously (Fig. 3a); with local connectivity, the outgoing connections only reach around the adjacent areas where the sampled neurons are located (Fig. 3b); and with uniform connectivity, the outgoing connections tend to cover the whole cortex evenly (Fig. 3c). Although the topology of three connectivity is different, they share the same out-degree distribution based on the way they are constructed (see Construction and storage of the connectivity).

To properly quantify the traveling waves observed in simulations, we applied a phase-based method from image processing Fleet and Jepson (1990), which has been adopted to analyze neuronal dynamics Rubino et al. (2006), and extended it to 3-D data. We first interpolate the simulated voltage onto a regular Cartesian grid that follows the Allen Brain Institute common coordinate framework (CCFv3); the voltage signal is then band-pass filtered into a narrow frequency band, and the Hilbert transform is applied to the narrowband-filtered grid signal to extract the band-specific generalized phase. We then compute the gradient field of the phase signal on the grid. The phase gradient directionality (PGD), an indicator of planar wave activity Rubino et al. (2006), is then used as our quantitative measure of macroscopic wave activity (see Quantitative measurement of neuronal activity for the full pipeline).

In Figure 4 and Videos, we compare the PGD produced by three representative simulations with different connectivity but the same parameters otherwise. The phase snapshots in Figure 4 span a 20 ms window centered on the PGD peak. Although both Allen and local connectivity produce visible wave activity, Allen connectivity produces waves that are much more macroscopic, while local connectivity produces locally organized planar waves, the form most frequently reported in previous experimental studies Lubenov and Siapas (2009); Bhattacharya et al. (2022) and computational models with local connectivity Keane and Gong (2015). Uniform connectivity does not generate any apparent structure. This qualitative observation is reflected in the PGD time series (Figure 4b; analogous comparisons in the delta, alpha, beta and gamma bands are shown in figure Supplement 2, figure Supplement 3, figure Supplement 4 and figure Supplement 5 respectively): Allen connectivity reaches the highest PGD in this representative simulation, with local and uniform remaining markedly lower.

Allen connectivity produces a higher level of macroscopic wave activity than local and uniform connectivity.

(a) Theta-band (4-8 Hz) phase snapshots from three representative simulations using Allen (top), local (middle) and uniform (bottom) connectivity under matched Poisson stimulation, sampled at −10, −5, 0, +5 and +10ms relative to the peak of phase gradient directionality (PGD). Colors represent the theta-band generalized phase of each neuron, normalized between 0 and 1. (b) Theta-band PGD of the three simulations versus time, aligned to the PGD peak. (c) Per-band comparison of the maximum PGD of the three representative simulations shown in (a), one bar per band per connectivity. Please see Videos for animations and Quantitative measurement of neuronal activity (illustrated in figure Supplement 1) for the calculation of PGD.

Network coupling and synchrony jointly shape macroscopic waves across the parameter plane

Traveling waves are known to propagate differently under different dynamical states, even when the underlying connectivity is fixed Muller et al. (2018). To assess whether this connectivity-dependent advantage holds beyond a single set of parameters, we systematically swept the coupling strength gAMPA and the Poisson stimulus magnitude Vstim on a 10 × 12 grid for each connectivity (Figure 5a, Simulation). Averaged over the full grid, Allen connectivity continues to produce the highest per-band PGD relative to local and uniform connectivity (Figure 5b), showing that the advantage seen Figure 4c is not specific to one set of parameters. We do note that for the delta band, local and Allen connectivity are able to generate a similar level of strong macroscopic wave activities. The same parameter plane also reveals three qualitatively distinct dynamic regimes for Allen connectivity (Figure 5a): coherent macroscopic waves at weak coupling, a disordered (asynchronous irregular) state at medium coupling in which wave structure collapses, and a globally synchronous regime at strong coupling that again supports waves.

Coupling strength, connectivity, and network synchrony jointly shape macroscopic-wave activity.

(a) Schematic of the three dynamic regimes that emerge across the (gAMPA, Vstim) parameter plane for Allen connectivity. At weak-to-intermediate coupling, the network supports coherent macroscopic traveling waves; at medium coupling, increased synaptic drive pushes the network into a disordered (asynchronous irregular) state where wave structure collapses; at strong coupling, the network reorganizes into a globally synchronous regime that again supports wave propagation. Insets show representative phase snapshots from each regime. (b) Per-band max PGD (mean ± SEM across the Poisson gAMPA × Vstim grid), plus the across-bands aggregate (‘All’), for Allen, local and uniform connectivity, with significance stars; full per-band (gAMPA, Vstim) heatmaps in figure Supplement 6. (c) Max PGD (alpha band) versus stimulus magnitude, error bars across gAMPA. (d) Max PGD (alpha band) versus excitatory conductance gAMPA, error bars across Vstim.(e) Kuramoto order parameter versus gAMPA in the alpha band. (f) Synchrony-PGD scatter, one point per (gAMPA, Vstim) combination, with Pearson r per connectivity (Allen r= 0.88, local r= 0.82, uniform r= 0.71 in alpha; all bands positive and significant at p < 10−3). Per-band versions of (c)-(e) are shown in figure Supplement 1, and per-band synchrony-PGD scatters in figure Supplement 2.

To dissect the non-monotonic coupling response that distinguishes the three regimes, we examine how PGD varies along each axis of the (gAMPA, Vstim) plane in the alpha band - the band that carries a large advantage for Allen connectivity (Figure 5b). Macroscopic wave activity is relatively insensitive to the stimulus magnitude (Figure 5c) but depends strongly and non-monotonically on the coupling strength (Figure 5d): for Allen connectivity, PGD peaks at weak coupling, collapses through a medium-coupling window (gAMPA ≈ 0.020-0.035 nS), and recovers at strong coupling. The uniform connectivity also follows this non-monotonic relationship, but has much lower max PGD at peaks compared to Allen connectivity. Local connectivity follows a different trajectory, peaking at weak coupling, then decaying almost monotonically with gAMPA. These trends hold band by band (figure Supplement 1): the peak-dip-recovery profile is visible in every band for Allen and uniform, while local decays almost monotonically across all bands.

To diagnose the mechanism behind this profile, we measured the Kuramoto order parameter R(t) = eiΦ(x,t) x from the generalized phase field Φ(x, t) of each frequency band and recorded its maximum over each simulation window (Quantitative measurement of neuronal activity). The resulting synchrony curve (Figure 5e) resembles the trend of the maximum PGD well (Figure 5d). For Allen connectivity in the alpha band, mean synchrony peaks at weak coupling, collapses in the medium-coupling window, and recovers at strong coupling. Across the full 120-point (gAMPA, Vstim) grid, synchrony and PGD are positively correlated in every band and every connectivity (Pearson r = 0.30-0.92, all p < 10−3; Figure 5f, with per-band scatters in figure Supplement 2). Therefore, the PGD trough at medium coupling may be a synchrony trough: increased synaptic drive pushes the network into an asynchronous irregular state, while strong coupling reestablishes rhythmic synchronization that supports wave propagation. This synchrony-to-wave bottleneck seems more significant to the networks with long-range connectivity (Allen, uniform), partly because long-range connections can augment synchrony in the coupled neuronal network Bazhenov et al. (2008). We also note that although uniform connectivity is able to achieve almost an identical level of synchrony to that of Allen connectivity, the PGD remains much lower due to the loss of spatial organization within.

The macroscopic wave findings extend to the local field potential estimate

Across the paper, our analyses use the intracellular voltage as the primary measurement for the PGD. To verify that the macroscopic wave structure we identify is also visible in a quantity comparable to experimentally recorded local field potentials (LFP), we followed a post-hoc pipeline Telenczuk et al. (2020) that reconstructs the net synaptic current received by every neuron directly from the voltage traces and the simulation’s connectivity (LFP estimation from intracellular voltage). The same 3-D phase-gradient pipeline is then applied to the reconstructed LFP. The propagating wavefront seen in the local-average voltage is recovered, with cleaner contrast, in the LFP estimate (Figure 6a; Videos). Voltage and LFP max PGD are tightly correlated across the parameter grid (Pearson r = 0.78-0.89 per band, p < 10−25 for every band; Figure 6b), and the LFP version carries roughly 2-3 × more PGD than the underlying voltage signal (paired p < 10−13 in every band; Figure 6c). Most importantly, Allen connectivity remains the highest-PGD connectivity in every band when measured on the LFP signal (Figure 6d), and the non-monotonic coupling profile from Figure 5 is preserved (Figure 6e). These results suggest that the macroscopic-wave findings are not specific to intracellular voltage and can also be recovered in an LFP-like signal.

Macroscopic waves are maintained by the local field potential estimate.

(a) Paired snapshots of the local-average voltage (top, clipped to [−65, −50] mV) and the synaptic-LFP estimate (bottom, z-scored) across five consecutive 5-ms frames in a representative Allen Poisson simulation. The same propagating wavefront sweeps across the cortex in both modalities. (b) Voltage versus LFP max PGD scatter for Allen connectivity, one point per (gAMPA, Vstim, band); Pearson r = 0.78-0.89 per band, all p < 10−25. (c) Mean max PGD per band, voltage versus LFP, for Allen connectivity (paired p < 10−13 for every band; LFP carries ≈ 2-3× higher PGD than the intracellular voltage). (d) LFP-based mean max PGD per band, three connectivities; Allen wins in every band against both local and uniform. (e) LFP max PGD versus gAMPA for Allen connectivity, one line per band; the same three-regime profile (peak, dip, recovery) seen in voltage is preserved in LFP. See LFP estimation from intracellular voltage for the LFP estimation pipeline.

Discussion

Construction of the Allen connectivity

In this work, we present an algorithm that integrates the spatial transcriptomic dataset, which provides the spatial distribution and molecular properties of neurons across the mouse brain, and the mesoscopic connectivity dataset that offers the connectivity information between brain regions on a voxelized level. The algorithm is efficient and can be easily adapted for similar datasets of the mouse brain or for even other species. In essence, our algorithm belongs to the family of stochastic block models Holland et al. (1983), where the block structure is given by the voxelization of the Allen Brain Atlas and the inter-block connection probabilities are set by the voxelized projection strengths. The projectome Knox et al. (2018) we use to build the Allen connectivity is derived from a series of viral-tracing experiments that reflect the connection strength between different regions Oh et al. (2014). However, the relationship between the actual connectome and the mesoscopic projectome remains unclear, and our current model inherits several important simplifications from this dataset.

First, the voxelized projection data is not cell-type specific. In our model, we distinguish only two neuronal populations: glutamatergic (excitatory) and GABAergic (inhibitory), based on the Zhuang-ABCA-1 transcriptomic dataset. However, real cortical connectivity differs dramatically between more refined cell types: pyramidal neurons and interneurons have distinct connection targets, and connectivity is strongly layer-specific. At the same time, we use a single global density parameter ρ to set the average number of connections per neuron across the entire cortex. A region-specific ρ could better capture the known variation in local synaptic density across cortical areas, for example the higher synaptic density of primary sensory regions relative to higher-order association areas. Future versions of the model could allow a region- and cell-type-specific ρ derived from region-level synapse-density atlases together with cell-type-resolved connectivity Harris et al. (2019), which would simultaneously address the limitations noted above.

Also, our model uses uniform synaptic weights within each synapse type: every AMPA synapse has conductance gAMPA and every GABA synapse has conductance gGABA. In reality, synaptic strength varies with presynaptic and postsynaptic cell type, anatomical distance, and the distribution of synaptic weights is typically heavy-tailed (e.g. log-normal Buzsáki and Mizuseki (2014)). Extending our algorithm to sample synaptic weights from realistic distributions would be a natural next step and is likely necessary for quantitative comparisons with electrophysiological recordings.

Cortical model and simulation

Compared to previous large-scale models Capone et al. (2023); Roberts et al. (2019), our model advances in incorporating the detailed spatial distribution of neurons in the 3-D computational domain. Without this information, spatiotemporal dynamics cannot be accurately analyzed and visualized. Recent research Pang et al. (2023); Gulbinaite et al. (2024) has highlighted the significant role that brain geometry plays in neuronal functions. In this work, we did not further use the spatial information of the cortex, such as the geometry of the cortical manifold, which would be an interesting direction for future work.

At the same time, although we are using one of the largest spatial transcriptomic datasets up to date, 300,000 neurons are still only a small portion of the total neuronal population in the mouse cortex. The number of neurons in a network will have important effects on aspects such as synaptic scaling. Among the total population, we only modeled two kinds of neurons, which is also a simplification. Therefore, enriching our model by including a higher number of different types of neurons and studying how they affect the conclusion in this paper is highly meaningful.

For simulations, we choose to randomly stimulate the total population of layer 4 neurons as a way to mimic subcortical input and generate traveling waves, which can be unrealistic. Subcortical structures, such as the thalamus, are vital to cortical dynamics like slow-wave activity Steriade et al. (1993) and are known to regulate traveling waves Ye et al. (2023). Therefore, a direct and important future improvement would be adding subcortical structures to the model. In this work, we choose two parameters that affect the dynamic regime of the network, which are the level of the stimuli controlled by the Poisson voltage bump and the coupling strength between neurons. Other parameters, including excitation-inhibition balance, are also likely to shape the dynamical regime Ahmadian and Miller (2021) and should be explored in future work.

Synaptic and conduction delays are essential to cortical dynamics such as traveling waves Petkoski and Jirsa (2022); Deco et al. (2009); Golomb and Ermentrout (2000). In our model, because the synapses are modeled by a first-order kinetic model (Neuronal and synaptic model), both excitatory and inhibitory synapses have a rising time that mimics the synaptic delay. However, we did not include conduction delay in our study. Conduction delay is thought to have a proportional relationship with the white matter path length of the connection between two neurons Lemaréchal et al. (2022). In our model, although we have the 3D positions of neurons, we do not have geo-metric information about the cortical manifold. Two neurons can be very close in the Cartesian coordinates measured by the Euclidean distance, but very far in terms of the length of the actual connection in the brain. Therefore, incorporating accurate conduction delay in the model is an important future direction.

Quantitative measurements of 3-D traveling waves

There are many techniques available for identifying and measuring large-scale neuronal spatiotem-poral patterns Townsend and Gong (2018); Gutzen et al. (2024). While methods for detecting wave dynamics in three-dimensional data do exist Budzinski et al. (2023), most published algorithms are designed to analyze 2-D data, so our simulation data, which is intrinsically 3-D, presents new challenges for measurement. For instance, we need to identify neuron boundaries to exclude un-realistic boundary effects when calculating the phase gradient field (Quantitative measurement of neuronal activity). Additionally, working with high-resolution 3-D data is computationally expensive. Despite these challenges, we extended and verified the previous phase-based method to quantify the level of macroscopic wave activity in our simulations. However, as noted in the original work Rubino et al. (2006), phase gradient directionality only indicates planar wave activity and may fail to identify other wave types such as spiral waves, which are observed in the cortex Muller et al. (2016). We also did not analyze other cortical dynamics, such as sinks and sources, which have appeared in our simulation. Extending more sophisticated methods from 2-D to 3-D to identify diverse spatiotemporal patterns in experimental or simulated data will be a meaningful direction for future research.

Comparing with experimental data

Across the paper, we have chosen the intracellular voltage of neurons as the primary data for the analysis. The voltage signals, though, do not directly resemble experimental measures such as local field potential (LFP), provide a quick way to analyze the overall dynamics of the whole network. There is strong experimental evidence of cortical spatiotemporal waves from measurements of LFP and multiunit activity reflecting extracellular electrical potentials in regions across the cortex Destexhe et al. (1999). Recent experimental studies of cortical wave patterns have achieved significant improvements in spatiotemporal resolution via methods such as wide-field calcium imaging Manley et al. (2024) and optical voltage imaging Liang et al. (2021). Our simulation framework allows many opportunities to compare with these experimental data.

However, one must be careful when comparing these simulations to experimental data, especially LFPs or other methods measuring extracellular potential rather than intracellular activity. Further work is needed to connect high-resolution mathematical models with experimental data based on the imaging techniques used. A more detailed quantitative comparison with experimental cortical-wave studies, such as the cortex-wide voltage-imaging data of Mohajerani et al. (2013) or the bidirectional visual-evoked waves reported by Aggarwal et al. (2022), is left as a future direction.

Neural oscillation and traveling wave

Different neural oscillations are known to have different generating mechanisms. For example, the gamma oscillation is conjectured to be caused by coordinated interaction of excitatory and inhibitory neurons in the cortex Buzsáki and Wang (2012). And slow-wave activity is conjectured to be caused by thalamocortical interaction Steriade et al. (1993). In our work, we did not fully investigate how different kinds of oscillations emerge in the network from realistic mechanisms, but only studied how these oscillations produce different levels of macroscopic traveling waves in the cortex a posteriori. In the future, it is essential to combine the established understanding of mechanisms behind different oscillations with experiments to further verify the conclusions drawn in this work.

Methods and Materials

Use of the spatial transcriptomic data

The original spatial transcriptomic study Zhang et al. (2023) provides four datasets of cell information identified from the mouse brain. Among these, we selected the largest dataset, Zhuang-ABCA-1, which includes 2.8 million cells from one hemisphere, with approximately 2.6 million (N = 2,616,328) cells registered to the Allen Brain Atlas. Of these 2.6 million cells, about half are non-neuronal cells and were removed before constructing the connectivity. Since we are only modeling the glutamatergic and GABAergic neurons, we only kept these two kinds of neuronal populations, totaling approximately 1 million neurons (N = 1,050,440). As a result, our final dataset consists of N = 1,050,440 neurons (Fig. 1a) in one hemisphere. Within this population, the isocortex contains 297,812 neurons. The cortical connectivity is constructed by applying our algorithm to the total cortical population.

Use of the voxelized connectivity data

The projection data we use in this study comes from Knox et al. (2018), which provides the connection strength between any pairwise voxel in the isocortex. The most important part of using the projection data is determining the number of connections between neurons in any pairwise voxels based on the connection strength. Let W represent the voxelized connectivity matrix, where Wab represents the connection strength from voxel a to voxel b. In this work, we assume the number of connections between two voxels is proportional to the connection strength, meaning the number of connections from voxel a to voxel b is Nab = λWab, where λ is a factor that will be calculated based on the density parameter.

Before building the connectivity, we first choose the density parameter, ρ, which is the average number of connections per neuron. The density parameter ρ is a free parameter that controls the trade-off between computational feasibility and biological fidelity: higher values of ρ yield more connections per neuron and thus higher biological realism, at the cost of greater memory and runtime Morrison et al. (2007). Let vector x represent the number of neurons in each voxel. If λ = 1, then the total number of connections in the whole connectivity is Ntotal=xTWx, and the average number of connections per neuron is N¯=xTWxM, where M is the total number of neurons. Therefore, if we let λ=ρN¯=ρ∗MxTWx, we will get the ideal average number of connections per neuron. Consequently, Nab=ρ∗MxTWxWab. In the code, after calculating λ beforehand, we can multiply the whole projection matrix W by λ to derive a connection matrix, whose each entry now represents the number of connections between any two voxels.

Construction and storage of the connectivity

Three types of connectivity are used in our simulations: Allen, local and uniform connectivity. Allen connectivity is constructed by applying our algorithm to the cortical neurons in Zhuang-ABCA1 based on the method described above. Uniform and local connectivity are constructed in a similar fashion: for each neuron, we first read out the number of outgoing connections in the corresponding Allen connectivity, then we randomly choose the same number of neurons from the total population without repetition for the uniform connectivity, or the same number of nearest neighbors for local connectivity. In this way, the out-degree distribution of all three connectivity will be the same, making them more comparable. In particular, local connectivity is constructed through nearest-neighbor coupling, where we use the K-nearest neighbors (KNN) algorithm Cover and Hart (1967) to choose a fixed number of outgoing connections to nearest neighbors for each neuron based on spatial position. For all three connectivity, self-connection is not allowed.

When running our connectivity algorithm to build the Allen connectivity, a large number of neurons can pose some challenges to our connectivity algorithm. To tackle this challenge, we loop through the voxels for a significant speed-up. As a result, our algorithm will take O(M N) time, where M is the number of neurons that contain all the neurons and N is the number of voxels. N has a maximum since the computational domain is finite, and in the end, this algorithm is about two orders of magnitude faster than iterating through every neuron.

In our model, we choose ρ = 300, which means there are about 90 million synapses in the model since we have 297,812 neurons. To store such huge connectivity, we use a compressed row form representation of the connectivity matrix, which is also used in similar packages Dai et al. (2020).

Neuronal and synaptic model

We use a Hodgkin-Huxley type neuron model that can produce different firing behaviors of a cortical neuron Pospischil et al. (2008). Each neuron has four ionic channels: voltage-gated sodium, voltage-gated potassium, slow non-inactivating potassium, and leak channels. The pyramidal neurons will have all four ionic channels, while the inhibitory neurons do not have the slow non-inactivating potassium channels because they are fast-firing and do not usually exhibit firing adaptation. Detailed equations of each current and the change of gating variables are provided in the appendix. The parameters of the neuronal model are also drawn from the original paper and presented in Table S1 in the appendix.

For synapses, we model both glutamatergic (AMPA) or GABAergic synapses in our cortical network. Both synapses are modeled by a conductance-based, first-order kinetic model. Please see the supplement for more details.

Simulation

To simulate the computational model, we have to solve two sets of equations at every time step: the Hodgkin-Huxley equations for the electrophysiology model and the conductance-coupled synaptic equations. We solve the Hodgkin-Huxley equations using an established implicit leapfrog method DeWoskin et al. (2015), allowing for large timesteps due to increased stability. Both electrophysiology and synaptic equations are solved in parallel on GPU in our framework. The overall parallelism is based on the neuron, which means each set of Hodgkin-Huxley equations of a neuron is put up on an individual thread of GPU to solve at every timestep. Similarly, at any time step, we first check for all neurons firing and then assign each thread to process each of the firing neurons’ downstream connections. Please see the supplements for a detailed description of the numerical methods.

Each simulation lasts for 3 seconds, and we only record the last 1 second of the simulation to avoid the transient at the beginning of the simulation. The initial conditions of all neurons are randomly chosen at the beginning of each simulation. To be more specific, the initial voltage of all neurons is drawn from a uniform distribution between −65 mV and −5 mV, and all gate variables are drawn from a uniform distribution between 0.02 and 0.05. For each simulation, we collect the voltage traces of all neurons with a sampling frequency of 2000 Hz.

In all simulations, the total excitatory population of layer 4 neurons is stimulated with a Poisson spike train at a rate of 10 Hz, delivering a fixed-amplitude voltage bump Vstim on each Poisson event. To adjust the level of drive we sweep Vstim from 5 to 32.5 mV in steps of 2.5 mV. To adjust the coupling strength we sweep the excitatory conductance gAMPA from 0.005 to 0.050 nS in steps of 0.005 nS, with the inhibitory conductance always fixed at 10% of the excitatory conductance. The full gAMPA ×Vstim grid (10 × 12 = 120 simulations per connectivity) is used for the Poisson-stimulation analyses in Figs. 4-6.

Quantitative measurement of neuronal activity

To quantify spatiotemporal dynamics, we extend a phase-based method Rubino et al. (2006) that has primarily been used for 2-D data but naturally generalizes to 3-D (figure Supplement 1 illustrates the full pipeline on a representative snapshot). Because the Hilbert transform is most interpretable when applied to narrowband signals, the simulated voltage traces are first interpolated onto a regular Cartesian grid that follows the Allen Brain Institute common coordinate framework (CCFv3), then band-pass-filtered into a frequency band, and only then converted to a generalized phase via the Hilbert transform. Concretely:

Spatial interpolation. We construct a regular 3-D Cartesian grid following the discretization of the Allen Brain Institute common coordinate framework (CCFv3), of size 132 × 80 × 114 at 100 μm spatial resolution, and linearly interpolate the per-neuron voltage time series onto this grid; grid cells that contain no neurons are masked out and excluded from later steps.

Band-pass filtering. The gridded voltage time series are bandpass-filtered into the five canonical bands (delta 0.5-4 Hz, theta 4-8 Hz, alpha 8-12 Hz, beta 12-30 Hz, gamma 30-100 Hz) using a fourth-order Butterworth filter applied with zero-phase filtfilt.

Generalized phase. For each narrowband-filtered gridded signal, we apply the Hilbert transform and extract the band-specific generalized phase Φ(x, t) for that band.

Phase gradient directionality. We compute the phase gradient ∇Φ on the 3-D grid using a finite-difference scheme and use it to compute the phase gradient directionality (PGD) P (t) Rubino et al. (2006), our quantitative measure of macroscopic wave activity:

(1) P(t)=‖∇ϕ―‖‖∇ϕ‖―, where the horizontal bar denotes the spatial average across the grid. Higher PGD means the gradient field is more aligned across spatial locations, indicating a more coherent planar wavefront.

In computing the gradient field, the boundary grid cells that contain no neurons are excluded by masking, implemented with the PyVista package Sullivan and Kaszynski (2019); both interpolation and gradient computation use the intrinsic routines of the 3-D Visualization Toolkit (VTK) Schroeder et al. (1998).

The Kuramoto order parameter R(t) = eiΦ(x,t) x is computed by averaging eiΦ across the valid (neuron-containing) grid cells of the same per-band phase field, and reduced to its maximum over each 1-s recording window for the synchrony curves in Figure 5.

The same per-band pipeline is run independently for each of the five bands in the per-band analyses (Figure 4c, Figure 5, Figure 6, figure Supplement 2-figure Supplement 5, figure Supplement 6, figure Supplement 1, figure Supplement 2). The LFP analyses in Figure 6 use the same pipeline applied to the per-neuron synaptic-LFP estimates (see LFP estimation from intracellular voltage) after interpolation onto the same grid.

LFP estimation from intracellular voltage

To connect our simulated intracellular voltage traces to experimentally measured local field potentials (LFP), we used a post-hoc pipeline that reconstructs the synaptic current received by every neuron directly from the voltage traces and the connectivity used in the simulation Telenczuk et al. (2020). Since LFP primarily reflects the net transmembrane synaptic current in a local population Destexhe et al. (1999), using the reconstructed synaptic current as an LFP proxy is a natural and widely adopted choice in large-scale simulations.

Concretely, for each neuron i at time t we first compute the sigmoidal neurotransmitter output used in the simulation’s synaptic model, (2) Ti(t)=11+exp⁡(−(Vi(t)−Vtc)/Kp), with Vtc = 2 mV and Kp = 5 mV matching the parameters used in the simulation. We then use the CSR connectivity matrix to propagate this presynaptic output to the postsynaptic neurons, keeping excitatory and inhibitory sources separate. Let AE and AI denote the incoming-connection adjacency matrices restricted to excitatory and inhibitory presynaptic neurons, respectively (obtained by transposing the outgoing CSR matrix and masking columns by the presynaptic E/I label); the excitatory and inhibitory synaptic activation at neuron j is (3) sjE(t)=∑iAjiETi(t),sjI(t)=∑iAjiITi(t). The net synaptic current per neuron is then (4) IjSyn(t)=gAMPAsjE(t)(Vj(t)−EAMPA)+gGABAsjI(t)(Vj(t)−EGABA), with reversal potentials EAMPA = 0 mV, EGABA = −70 mV, and the same conductances gAMPA, gGABA used in the corresponding simulation. This per-neuron current Isyn(t) plays the role of an LFP-like proxy: spatially averaging Isyn over a local population yields a signal analogous to an extracellularly measured LFP. The pipeline is implemented with sparse matrix multiplications and streams the voltage traces in chunks, so it scales to the full 297,812-neuron population across the 1-second recording window.

We then run the same 3-D phase-gradient pipeline described in Quantitative measurement of neuronal activity on the reconstructed LFP. The per-neuron synaptic current is interpolated onto the Allen Brain Atlas 3-D grid, bandpass-filtered into the delta, theta, alpha, beta, and gamma bands, converted to a phase field via the Hilbert transform, and used to compute PGD.

Visualization

Throughout the paper, we visualize neurons by coloring them based on their intracellular voltage. The visualization is straightforward for a single neuron such as in figure Supplement 1, but in Figure 2, Figure 4 and Figure 6, we also need to calculate the average voltage of 300 neurons nearby. This is done by identifying the closest 300 neighbors for each neuron using the K-Nearest-Neighbor algorithm and calculating the average. All visualizations are made using the PyVista package Sulli-van and Kaszynski (2019).